### 2015-09-30 paper/paper-refsq2016-submission.tex
add data files

%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is the LaTeX source for the instructions to authors using
% the LaTeX document class 'llncs.cls' for contributions to
% the Lecture Notes in Computer Sciences series.
% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
%
% It may be used as a template for your own input - copy it
% to a new file with a new name and use it as the basis
% for your article.
%
% NB: the document class 'llncs' has its own and detailed documentation, see
% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\modelname, \modelname

%\documentclass[runningheads,a4paper]{llncs}
\documentclass[a4paper]{llncs}

\usepackage{tikz}
\usetikzlibrary{positioning,fit}

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
%%%%TODO: REMOVE THE TERM metacompleteness
%% MY COOL PACKAGES
\usepackage{times}
\usepackage[normalem]{ulem}
\usepackage{cite}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{scalefnt}
\usepackage{setspace}
%
%\usepackage{listings}
%\lstset{basicstyle=\ttfamily\small, 
%%  numbers=left, numberstyle=\tiny, stepnumber=1, numbersep=5pt,
%  language=java, mathescape=true, escapechar=\!,
%  emph={object}, emphstyle=\textbf,
%  showstringspaces=false
%}
%%%%

\usepackage{url}
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

\begin{document}

%\floatstyle{ruled}
%\newfloat{Specification}{htbp}{lop}

\mainmatter  % start of an individual contribution

% first the title is needed
\title{reqT.org --  Towards a Semi-Formal, Open and Scalable Requirements Modeling Tool}
%reqT --  A Requirements Modeling Tool for Code Lovers
% a short form should be given in case it is too long for the running head
%\titlerunning{DRAFT MANUSCRIPT - Please do not cite. Submitted to REFSQ'13.}

\author{Bj\"orn Regnell}
%
%\authorrunning{Regnell, "Requirements Modeling for Code Lovers"}
\institute{Dept. of Computer Science, Lund University, Sweden \\ \url{bjorn.regnell@cs.lth.se} }

\maketitle

%%%% ABSTRACT
\begin{abstract}
[{\bf Context and motivation}] This research preview presents ongoing work on a free software requirements modeling tool called reqT that is developed in an educational context. [{\bf Question/problem}] The work aims to engage computer science students in Requirements Engineering (RE) through a tool that captures essential RE concepts in executable code. [{\bf Principal ideas}] Requirements are modeled using an internal DSL in the Scala programming language that blends natural language strings with a graph-oriented formalism. [{\bf Contribution}] The metamodel of reqT and its main features are presented and modeling examples are provided together with a discussion on initial experiences from student projects, limitations and directions of further research.   
\keywords{requirements engineering, requirements modeling, software engineering, CASE tool, requirements metamodel, requirements engineering education, internal DSL, embedded DSL, Scala programming language}
\end{abstract}

%%%%% INTRO
\section{Introduction}
There are many challenges in teaching Requirements Engineering (RE) \cite{Memon2010, Regev2011}, including  conveying requirements modeling skills \cite{Callele2006}. Given a wide-spread attention on agile methods with less emphasis on extra-code artifacts \cite{Ramesh2010}, it may be particularly challenging to motivate coding-focused engineering students (and software practitioners) to spend serious effort on requirements modeling. One way to inspire software engineers to learn more about and do more RE may be to offer an interesting software tool. There are nowadays numerous commercial RE tools available, but many are expensive, complex and not sufficiently open  \cite{Carillo2011}.

This paper presents on-going work on a tool named reqT that aims to provide a small but scalable, semi-formal and free software package for an educational setting that (hopefully) can inspire code lovers to learn more about requirements modeling. A long-term goal of reqT is to offer an open platform for RE research prototypes, e.g. for feature modeling and release planning research. The tool development started in 2011 at Lund University, where reqT is used in RE teaching at MSc level in the Computer Science \& Engineering program. In 2012 reqT was rebuilt from scratch based on student feedback. The tool can be downloaded from: \url{http://reqT.org}

The paper is organized as follows. Section \ref{section:objectives} states the objectives and motivates the design strategy of reqT. Section \ref{section:metamodel} presents the metamodel of reqT and some example reqT models. Section \ref{section:discussion} discusses limitations and some initial experiences from using reqT in teaching and concludes the paper with a sketch of future research directions.

\section{Goals, Design Strategy and Rationale}\label{section:objectives}
The main objective behind reqT is to establish a set of essential RE concepts and capture them in an expressive, extensible and executable language appealing to computer science students (and eventually practitioners). This general objective is accompanied by the following main goals and associated design strategies:
\begin{enumerate}
\item {\bf Semi-formal.} {\it Goal:} Provide a semi-formal representation of typical requirements modeling constructs that can illustrate a flexible combination of expressive natural language-style requirements with type-safe formalisms allowing static checks. {\it Design:} Use graph structures based on typed nodes representing typical requirement entities and attributes, and typed edges representing typical requirements relations, and implement the graph as an associative array (map). {\it Why?} Graphs are well-known to many CS students. Maps are efficient from an implementation perspective and may be less complex to master compared to e.g. SQL databases.
\item {\bf Open.} {\it Goal:} Provide a platform-independent requirements tool that is free of charge. {\it Design:} Use Java Virtual Machine technology and release the code under an open source license. Use tab-separated, tabular text-files for import and export. Use HTML for document generation. {\it Why?} There are many free libraries available that runs on a JVM. Tab-sep and HTML support interoperability. %A free, platform-independent tool is appreciated by many students.
\item {\bf Scalable.} {\it Goal:}  Provide an extensible requirements modeling language that can scale from small, concise models to large families of models with thousands of requirements entities and relations. {\it Design:} Implement reqT as an internal DSL (Domain-Specific Language) in the Scala programming language \cite{Odersky2004}. Use Map and Set from Scala collections to represent requirements graphs. {\it Why?} Scala is a modern, statically typed language with an efficient collections library. Scala offers scripting abilities that provide general extensibility without re-compilation. Integrated development environments \cite{eclipse}, as well as interactive scripting tools are available \cite{kojo}. %When requirements are expressed as code, general software engineering tools can be applied when scaling to large projects, e.g. configuration management tools and cloud-based code repositories to facilitate requirements evolution and collaboration over Internet. 
\end{enumerate}

These goals, design strategies and rationale are directing the on-going work, and it remains to be investigated to what extent the main objective and goals can be met. A critical issue is how to interpret what are "essential" RE concepts and "typical" modeling constructs. The reqT tool is used in a course based on a specific text book \cite{Lauesen2002} and a specific student project concept \cite{ets170}, and the concepts of the reqT requirements metamodel (see Fig. \ref{fig:types}) reflect that context. However, the reqT architecture is prepared for extensions of new concepts in the metamodel to cater for different educational contexts. 
%%%%%%% THE MODEL
\section{Modeling requirements with reqT}\label{section:metamodel}

A reqT model includes sequences of graph parts {\tt<Entity><Edge><NodeSet>}  separated by comma and wrapped inside a {\tt Model( )} construct. A small reqT Model with three {\tt Feature} entities and one {\tt Stakeholder} entity is shown below:
\begin{spacing}{0.8}
\begingroup
    \fontsize{8pt}{12pt}\selectfont
\begin{verbatim}
Model(
  Feature("f1") has (Spec("A good spec."), Status(SPECIFIED)),
  Feature("f1") requires (Feature("f2"), Feature("f3")),
  Stakeholder("s1") assigns(Prio(1)) to Feature("f2")
)
\end{verbatim}
\endgroup
\end{spacing}

%%%%%%%%%%%%%% DISCUSSION

\section{Discussion and Conclusion}\label{section:discussion}

 The results of the on-going work with reqT remains to be further investigated and a validation of reqT as a RE learning tool and research experimentation platform is subject to future work. This section discusses some preliminary experiences, limitations, relation to state-of-the-art and future research directions.

{\bf Preliminary proof-of-concept.}
The first version of reqT was tried on a voluntary basis by 12 students working in groups of 6 students each during fall 2011. Statements from course evaluations indicate that the students found reqT useful in their learning. One group used a configuration management tool for reqT models to manage  their parallel work, while one group used a cloud service and tab-sep export/import to collaborate over the Internet. The group with the largest requirements model produced 64 features, 18 tasks, 12 functions, 30 data requirements and 33 quality requirements, in total 157 requirements entities.

Several students appreciated that reqT can mix informal text with a graph-oriented formalism, but some requested more elaborated functionality for document generation, as well as linking to external images. Some students also requested more modeling examples that show how the text book techniques could be transferred to reqT models.

Based on student feedback, reqT was rebuilt from scratch during 2012 with a new architecture and a new version of the meta model (see Fig. \ref{fig:types}), as well as a revised Scala-internal DSL. The template-controlled HTML generation was implemented based on student suggestions. The teaching material was complemented with more example models directly related to the textbook. The second version of reqT is currently tested by  students in a new course instance and a post-course evaluation of reqT is planned in spring 2013.

Our preliminary experiences from applying reqT in teaching suggest that reqT, if used in a suitable teaching context, may encourage students with a code-focused mind set to learn and practice RE in the following ways: (1) A free and platform-independent software tool that is implemented using a modern programming language with interactive scripting facilities can attracts the interest of code-focused students. (2) Requirements can be processed, queried, transformed or exported using Scala scripts, and the open-ended nature of reqT that allows students to code their own scripts to both manage requirements models and to adapt reqT to fit their RE needs in the course project was appreciated by several coding-literate students. (3) By turning requirements models into executable code, students can use programming tools such as a console command line interpreter (the Scala REPL) as well as a source code version control system (e.g. git-scm.com) to branch and merge their collaborative work on requirements in ways they are used to from their previous collaborative software implementation courses, including issue tracking systems and code review support.

{\bf Relation to state-of-the-art.} To the best of our knowledge there is no other RE tool that allows semi-formal requirement models to become executable programs through an internal Scala DSL, and thus letting coding, testing and requirements engineering share the same media. In the RE tool survey by Carrillo de Gea et al.  \cite{Carillo2011} it is pointed out that "many expensive tools aren't sufficiently open". The reqT technology aims to be completely free and open to facilitate academic usage, collaborative evolution and incorporation of new RE concepts in different teaching and research contexts. Many of the existing tools have proprietary representations \cite{Carillo2011}, while users of reqT can extend the reqT metamodel with new entities and attributes simply by adding case classes with a few lines of code. However, reqT cannot compete with versatile commercial RE tools  \cite{Carillo2011}  in terms of e.g. features completeness and graphical user interface.

{\bf Limitations. } In its current version, reqT has a number of limitations: (1) As the user interface is text based and depends on the command line interface of the Scala REPL or a script editor environment \cite{kojo, eclipse}, students that only are prepared to use graphical user interfaces may be discouraged. Some of our students preferred to work in a GUI spreadsheet application using tab-separated exports from reqT that was generated by other team members assigned by the student group to be reqT experts. (2) It requires some knowledge of Scala to tailor reqT exports and there is a need for a more comprehensive API for adaptable document generation. (3) The embedded DSL requires some learning efforts and it remains to be investigated if the effort is justified by the knowledge gained. (4) To support scalability to large families of reqT models there is a need for modularization concepts and overview visualizations. (5) The explicit typing of entities with keywords such Feature and Stakeholder can be perceived as verbose compared to more concise but potentially cryptic abbreviations (e.g. Fe, Sh). This may be addressed by DSL-specific editor support, such as code-completion, code folding and code templates.

{\bf Future work.} Further directions of research include (1) incorporation of constraints on models for support of prioritization and release planning \cite{Regnell2011}, (2) more elaborate semantic checks to better guide requirements modelers, and (3) graphical visualization of requirements graph models. (4) Natural Language Processing technology including e.g. ambiguity risk detection may be interesting to combine with reqT. (5) It is also important to further investigate the pedagogic advantages and limitations of the approach.

A major objective of this research preview paper is to expose the latest version of reqT to the community of RE scholars and to invite discussions and contributions.

\bibliography{refs}
\bibliographystyle{splncs03}

\clearpage
\hspace{-0.5cm}
\begin{minipage}{1\linewidth}
\begin{minipage}{0.36\textwidth} 
\begin{table}[H]
%\centering
\vspace{-0.7cm}
\fontsize{6}{6.5}\selectfont
\caption{Definitions of concepts.}%: Entities, Attributes and Relations.}
\label{table-entities}
\begin{tabular}{|l p{5.0cm}|}
\hline
\textit{Entity} & \textit{Definition}  \\ \hline
Actor&A human or machine that communicates with a system.\\
App&A computer program, or group of programs designed for end users, normally with a graphical user interface. Short for application.\\
Barrier&Something that makes it difficult to achieve a goal or a higher quality level.\\
Breakpoint&A point of change. An important aspect of a (non-linear) relation between quality and benefit.\\
Class&An extensible template for creating objects. A set of objects with certain attributes in common. A category.\\
Component&A composable part of a system. A reusable, interchangeable system unit or functionality.\\
Configuration&A specific combination of variants.\\
Data&Information stored in a system.\\
Design&A specific realization or high-level implementation description (of a system part).\\
Domain&The application area of a product with its surrounding entities.\\
Epic&A large user story or a collection of stories.\\
Event&Something that can happen in the domain and/or in the system.\\
Feature&A releasable characteristic of a product. A (high-level, coherent) bundle of requirements.\\
Function&A description of how input data is mapped to output data. A capability of a system to do something specific.\\
Goal&An intention of a stakeholder or desired system property.\\
Idea&A concept or thought (potentially interesting).\\
Interface&A defined way to interact with a system.\\
Issue&Something needed to be fixed.\\
Item& An article in a collection, enumeration, or series.\\
Label&A descriptive name used to identify something.\\
Member&An entity that is part of another entity, eg. a field in a in a class.\\
Meta&A prefix used on a concept to mean beyond or about its own concept, e.g. metadata is data about data.\\
MockUp&A prototype with limited functionality used to demonstrate a design idea.\\
Module&A collection of coherent functions and interfaces.\\
Product&Something offered to a market.\\
Quality&A distinguishing characteristic or degree of goodness.\\
Relationship&A specific way that entities are connected.\\
Release&A specific version of a system offered at a specific time to end users.\\
Req&Something needed or wanted. An abstract term denoting any type of information relevant to the (specification of) intentions behind system development. Short for requirement.\\
Resource&A capability of, or support for development.\\
Risk&Something negative that may happen.\\
Scenario&A (vivid) description of a (possible future) system usage.\\
Screen&A design of (a part of) a user interface.\\
Section&A part of a (requirements) document.\\
Service&Actions performed by systems and/or humans to provide results to stakeholders.\\
Stakeholder&Someone with a stake in the system development or usage.\\
State&A mode or condition of something in the domain and/or in the system. A configuration of data.\\
Story&A short description of what a user does or needs. Short for user story.\\
System&A set of interacting software and/or hardware components.\\
Target&A desired quality level or goal .\\
Task&A piece of work (that users do, maybe supported by a system).\\
Term&A word or group of words having a particular meaning.\\
Test&A procedure to check if requirements are met.\\
Ticket&(Development) work awaiting to be completed.\\
UseCase&A list of steps defining interactions between actors and a system to achieve a goal.\\
User&A human interacting with a system.\\
Variant&An object or system property that can be chosen from a set of options.\\
VariationPoint&An opportunity of choice among variants.\\
WorkPackage&A collection of (development) work tasks.\\
\hline
 \end{tabular}
\end{table}
\end{minipage}
 \hspace{4em plus 1fill} 
\begin{minipage}{0.47\textwidth} 
 \begin{table}[H]
%\centering
\fontsize{6}{6.5}\selectfont
%\caption{Definitions of attributes and relations.}
\vspace{-0.2cm}
\label{table-entities}
\begin{tabular}{|l p{4.3cm}|}
 \hline
\textit{Attribute} & \textit{Definition}  \\ \hline
Benefit&A characterisation of a good or helpful result or effect (e.g. of a feature).\\
Capacity&The largest amount that can be held or contained (e.g. by a resource).\\
Code&A collection of (textual) computer instructions in some programming language, e.g. Scala. Short for source code.\\
Comment&A note that explains or discusses some entity.\\
Constraints&A collection of propositions that restrict the possible values of a set of variables.\\
Cost&The expenditure of something, such as time or effort, necessary for the implementation of an entity.\\
Damage&A characterisation of the negative consequences if some entity (e.g. a risk) occurs.\\
Deprecated&A description of why an entity should be avoided, often because it is superseded by another entity, as indicated by a 'deprecates' relation.\\
Example&A note that illustrates some entity by a  typical instance.\\
Expectation&The required output of a test in order to be counted as passed.\\
FileName&The name of a storage of serialized, persistent data.\\
Frequency&The rate of occurrence of some entity. \\
Gist&A short and simple description of an entity, e.g. a function or a test.\\
Image&(The name of) a picture of an entity.\\
Input&Data consumed by an entity, \\
Max&The maximum estimated or assigned (relative) value.\\
Min&The minimum estimated or assigned (relative) value.\\
Order&The ordinal number of an entity (1st, 2nd, ...).\\
Output&Data produced by an entity, e.g. a function or a test.\\
Prio&The level of importance of an entity. Short for priority.\\
Probability&The likelihood that something (e.g. a risk) occurs.\\
Profit&The gain or return of some entity, e.g. in monetary terms.\\
Spec&A (detailed) definition of an entity. Short for specification\\
Status&A level of refinement of an entity (e.g. a feature) in the development process. \\
Text&A sequence of words (in natural language).\\
Title&A general or descriptive heading.\\
Value&An amount. An estimate of worth.\\
Why&A description of intention. Rationale.\\

\hline
\textit{Relation} & \textit{Definition}  \\ \hline
binds&Ties a value to an option. A configuration binds a variation point.\\
deprecates&Makes outdated. An entity deprecates (supersedes) another entity.\\
excludes&Prevents a combination. An entity excludes another entity.\\
has&Expresses containment, substructure. An entity contains another entity.\\
helps&Positive influence. A goal helps to fulfil another goal.\\
hurts&Negative influence. A goal hinders another goal.\\
impacts&Some influence. A new feature impacts an existing component.\\
implements&Realisation of. A module implements a feature.\\
interactsWith&Communication. A user interacts with an interface.\\
is&Sub-typing, specialization, part of another, more general entity.\\
precedes&Temporal ordering. A feature precedes (is implemented before) another feature.\\
relatesTo&General relation. An entity is related to another entity.\\
requires&Requested combination. An entity is required (or wished) by another entity.\\
superOf&Super-typing, generalization, includes another, more specific entity.\\
verifies&Gives evidence of correctness. A test verifies the implementation of a feature.\\
 \hline
\end{tabular}
\end{table}
\end{minipage}
\end{minipage}
\subsubsection*{Acknowledgments.} This work is partly funded by  VINNOVA within the EASE project.

\end{document}

### 2015-10-01 paper/paper-refsq2016-submission.tex
update paper

@@
 % first the title is needed
-\title{reqT.org --  Towards a Semi-Formal, Open and Scalable Requirements Modeling Tool}
+\title{What is essential? -- A pilot survey on views about the requirements metamodel of reqT}
 %reqT --  A Requirements Modeling Tool for Code Lovers
@@
 \begin{abstract}
-[{\bf Context and motivation}] This research preview presents ongoing work on a free software requirements modeling tool called reqT that is developed in an educational context.
-[{\bf Question/problem}] The work aims to engage computer science students in Requirements Engineering (RE) through a tool that captures essential RE concepts in executable code.
-[{\bf Principal ideas}] Requirements are modeled using an internal DSL in the Scala programming language that blends natural language strings with a graph-oriented formalism.
-[{\bf Contribution}] The metamodel of reqT and its main features are presented and modeling examples are provided together with a discussion on initial experiences from student projects, limitations and directions of further research.
-\keywords{requirements engineering, requirements modeling, software engineering, CASE tool, requirements metamodel, requirements engineering education, internal DSL, embedded DSL, Scala programming language}
+[{\bf Context and motivation}] This research preview presents ongoing work on the metamodel of a free software requirements modeling tool called reqT that is developed in an educational context.
+[{\bf Question/problem}] The work aims to make an initial validation of a survey instrument that elicits views on the metamodel of the reqT tool, which seek to engage computer science students in Requirements Engineering (RE) through an open source requirements engineering DSL embedded in the Scala programming language.
+The research question underpinning the presented work is: Which RE concepts are essential to include in the metamodel for a requirements engineering tool in an educational context?
+[{\bf Principal ideas}] A survey instrument is developed with a list of 92 concepts (49 entities, 15 relations and 28 attributes) and a set of questions for each concept that elicit the respondents' views on the usage and interpretation of each concept.
+[{\bf Contribution}] The survey is initially validated in a pilot study involving 14 Swedish RE scholars as subjects.
+The survey results indicate that the survey is feasible if the respondents is willing to invest around 30 minutes of their time.
+The analysis of the responses suggest that many of the concepts in the metamodel are used frequently by the respondents and there is a large degree of agreement among the respondents about the meaning of the concepts.
+The results are encouraging for future work on empirical validation of the relevance of the reqT metamodel.
+
+\keywords{requirements engineering, requirements metamodel, CASE tool, requirements engineering education, embedded domain-specific language, empirical software engineering}
 \end{abstract}

### 2015-10-01 run.sh
add script with runners for bash/powershell

@@ -1,0 +1,1 @@
+scala -cp "./lib/poi-3.12/poi-3.12-20150511.jar:./lib/poi-3.12/poi-ooxml-3.12-20150511.jar" src/script.scala

### 2015-10-01 src/script.scala

@@ -1,0 +1,157 @@
+/* run with:
+scala -cp "./lib/poi-3.12/poi-3.12-20150511.jar;./lib/poi-3.12/poi-ooxml-3.12-20150511.jar" src/script.scala
+*/
+import org.apache.poi.poifs.filesystem.POIFSFileSystem
+import org.apache.poi.ss.usermodel.WorkbookFactory
+import org.apache.poi.ss.usermodel.Row
+import org.apache.poi.ss.usermodel.Cell
+import org.apache.poi.ss.usermodel.DataFormatter
+import java.io._
+import scala.collection.JavaConversions._
+import scala.util.Try
+
+case class Matrix[T](vectors: Vector[Vector[T]]){
+  val nrows = vectors.size
+  lazy val ncols = vectors.map(_.size).max
+  lazy val dim = (nrows, ncols)
+  def row(r: Int): Vector[T] = vectors(r)
+  def col(c: Int): Vector[T] = vectors.map(row => row(c))
+  lazy val transpose = Matrix((0 until ncols).map(c => col(c)).toVector)
+  def apply(x: Int, y: Int): T = vectors(x)(y)
+  def get(x: Int, y: Int): Option[T] = Try(vectors(x)(y)).toOption
+}
+
+object Xls {
+
+  def load(fileName: String, sheetNum: Int = 0): Matrix[String] = 
+    loadFile(new File(fileName), sheetNum)
+
+  def loadFile(file: File, sheetNum: Int = 0): Matrix[String] = {
+    val wb = WorkbookFactory.create(file)
+    val sheet = wb.getSheetAt(sheetNum); 
+    val nrows = sheet.getLastRowNum
+    val rowsMaybeNull: Vector[Row]  = (0 until nrows).map(r => sheet.getRow(r)).toVector
+    val ncols = rowsMaybeNull.map(r => Option(r).map(_.getLastCellNum).getOrElse(0.toShort)).max
+    val cellMatrix: Vector[Vector[Cell]] = rowsMaybeNull.map(r => 
+      if (r != null) (0 until ncols).map(c => r.getCell(c)).toVector else Vector.fill(ncols)(null))
+    val vectors: Vector[Vector[String]] = 
+      cellMatrix.map(_.map(cell => cellToString(cell)))
+    Matrix(vectors)
+  } 
+
+  lazy val dataFormatter = new DataFormatter(new java.util.Locale("en"))
+
+  def cellToString(cell: Cell): String = 
[…]

### 2015-10-01 analysis/entity-defs.tex
add analysis and update paper

\texttt{Actor}&A human or machine that communicates with a system.\\
\texttt{App}&A computer program, or group of programs designed for end users, normally with a graphical user interface. Short for application.\\
\texttt{Barrier}&Something that makes it difficult to achieve a goal or a higher quality level.\\
\texttt{Breakpoint}&A point of change. An important aspect of a (non-linear) relation between quality and benefit.\\
\texttt{Class}&An extensible template for creating objects. A set of objects with certain attributes in common. A category.\\
\texttt{Component}&A composable part of a system. A reusable, interchangeable system unit or functionality.\\
\texttt{Configuration}&A specific combination of variants.\\
\texttt{Data}&Information stored in a system.\\
\texttt{Design}&A specific realization or high-level implementation description (of a system part).\\
\texttt{Domain}&The application area of a product with its surrounding entities.\\
\texttt{Epic}&A large user story or a collection of stories.\\
\texttt{Event}&Something that can happen in the domain and/or in the system.\\
\texttt{Feature}&A releasable characteristic of a product. A (high-level, coherent) bundle of requirements.\\
\texttt{Function}&A description of how input data is mapped to output data. A capability of a system to do something specific.\\
\texttt{Goal}&An intention of a stakeholder or desired system property.\\
\texttt{Idea}&A concept or thought (potentially interesting).\\
\texttt{Interface}&A defined way to interact with a system.\\
\texttt{Issue}&Something needed to be fixed.\\
\texttt{Item}& An article in a collection, enumeration, or series.\\
\texttt{Label}&A descriptive name used to identify something.\\
\texttt{Member}&An entity that is part of another entity, eg. a field in a in a class.\\
\texttt{Meta}&A prefix used on a concept to mean beyond or about its own concept, e.g. metadata is data about data.\\
\texttt{MockUp}&A prototype with limited functionality used to demonstrate a design idea.\\
\texttt{Module}&A collection of coherent functions and interfaces.\\
\texttt{Product}&Something offered to a market.\\
\texttt{Quality}&A distinguishing characteristic or degree of goodness.\\
\texttt{Relationship}&A specific way that entities are connected.\\
\texttt{Release}&A specific version of a system offered at a specific time to end users.\\
\texttt{Req}&Something needed or wanted. An abstract term denoting any type of information relevant to the (specification of) intentions behind system development. Short for requirement.\\
\texttt{Resource}&A capability of, or support for development.\\
\texttt{Risk}&Something negative that may happen.\\
\texttt{Scenario}&A (vivid) description of a (possible future) system usage.\\
\texttt{Screen}&A design of (a part of) a user interface.\\
\texttt{Section}&A part of a (requirements) document.\\
\texttt{Service}&Actions performed by systems and/or humans to provide results to stakeholders.\\
\texttt{Stakeholder}&Someone with a stake in the system development or usage.\\
\texttt{State}&A mode or condition of something in the domain and/or in the system. A configuration of data.\\
\texttt{Story}&A short description of what a user does or needs. Short for user story.\\
\texttt{System}&A set of interacting software and/or hardware components.\\
\texttt{Target}&A desired quality level or goal .\\
\texttt{Task}&A piece of work (that users do, maybe supported by a system).\\
\texttt{Term}&A word or group of words having a particular meaning.\\
\texttt{Test}&A procedure to check if requirements are met.\\
\texttt{Ticket}&(Development) work awaiting to be completed.\\
\texttt{UseCase}&A list of steps defining interactions between actors and a system to achieve a goal.\\
\texttt{User}&A human interacting with a system.\\
\texttt{Variant}&An object or system property that can be chosen from a set of options.\\
\texttt{VariationPoint}&An opportunity of choice among variants.\\
\texttt{WorkPackage}&A collection of (development) work tasks.\\

### 2015-10-01 analysis/essential.tex
add analysis and update paper

%%%%%%%%%%%%% Summary table of (use, agree) >= (1, 2)
$14$ & \texttt{Class, Component, UseCase, Variant} & \texttt{Comment, Example, Max, Min, Title} & \texttt{implements, verifies}\\\hline
$13$ & \texttt{Configuration, Data, Design, Event, Quality, Scenario, Stakeholder, System, Term} & \texttt{Code, Constraints, Cost, FileName, Probability, Profit, Spec, Why} & \texttt{excludes, interactsWith, is, relatesTo, requires}\\\hline
$12$ & \texttt{Actor, Domain, Feature, Function, Interface, Module, Relationship, Release, Req, Risk, Service, State, Task, Test} & \texttt{Benefit, Capacity, Frequency, Input, Order, Output, Prio, Text, Value} & \texttt{has, impacts}\\\hline
$11$ & \texttt{Idea, Label, Member, Meta, MockUp, Section, User} & \texttt{Image} & \texttt{precedes, superOf}\\\hline
$10$ & \texttt{Goal, Story} & \texttt{Expectation} & \texttt{}\\\hline
$9$ & \texttt{App, Issue, Target, WorkPackage} & \texttt{Damage} & \texttt{binds, helps}\\\hline
$8$ & \texttt{Item, Product, Resource, VariationPoint} & \texttt{} & \texttt{deprecates}\\\hline
$7$ & \texttt{Breakpoint, Screen} & \texttt{Status} & \texttt{}\\\hline
$6$ & \texttt{Barrier} & \texttt{Deprecated} & \texttt{hurts}\\\hline
$5$ & \texttt{} & \texttt{} & \texttt{}\\\hline
$4$ & \texttt{Ticket} & \texttt{} & \texttt{}\\\hline
$3$ & \texttt{} & \texttt{} & \texttt{}\\\hline
$2$ & \texttt{} & \texttt{} & \texttt{}\\\hline
$1$ & \texttt{Epic} & \texttt{Gist} & \texttt{}\\\hline
$0$ & \texttt{} & \texttt{} & \texttt{}\\\hline

### 2015-10-01 analysis/relation-defs.tex
add analysis and update paper

\texttt{binds}&Ties a value to an option. A configuration binds a variation point.\\
\texttt{deprecates}&Makes outdated. An entity deprecates (supersedes) another entity.\\
\texttt{excludes}&Prevents a combination. An entity excludes another entity.\\
\texttt{has}&Expresses containment, substructure. An entity contains another entity.\\
\texttt{helps}&Positive influence. A goal helps to fulfil another goal.\\
\texttt{hurts}&Negative influence. A goal hinders another goal.\\
\texttt{impacts}&Some influence. A new feature impacts an existing component.\\
\texttt{implements}&Realisation of. A module implements a feature.\\
\texttt{interactsWith}&Communication. A user interacts with an interface.\\
\texttt{is}&Sub-typing, specialization, part of another, more general entity.\\
\texttt{precedes}&Temporal ordering. A feature precedes (is implemented before) another feature.\\
\texttt{relatesTo}&General relation. An entity is related to another entity.\\
\texttt{requires}&Requested combination. An entity is required (or wished) by another entity.\\
\texttt{superOf}&Super-typing, generalization, includes another, more specific entity.\\
\texttt{verifies}&Gives evidence of correctness. A test verifies the implementation of a feature.\\

### 2015-10-01 paper/paper-refsq2016-submission.tex
add analysis and update paper

@@
 %% MY COOL PACKAGES
-\usepackage{times}
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+%\usepackage{times}
+\usepackage{tgtermes}
+%\usepackage{bera}
+\usepackage[scaled=0.93]{beramono}
 \usepackage[normalem]{ulem}
@@
 
-\usepackage{url}
+\usepackage{hyperref}
+\usepackage{xcolor}
+\definecolor{entityColor}{RGB}{0,100,200}
+\definecolor{attributeColor}{RGB}{0,100,50}
+\definecolor{relationColor}{RGB}{160,0,30}
+\usepackage{listings}
+\lstdefinestyle{reqT}{
+  %belowcaptionskip=1\baselineskip,
+  breaklines=true,
+  %showstringspaces=false,
+  showspaces=false,
+  %breakatwhitespace=true,
+  basicstyle=\ttfamily\scriptsize,
+  emph={Ent,Meta,Item,Label,Section,Term,Actor,App,Component,Domain,Module,Product,Release,Resource,Risk,Service,Stakeholder,System,User,Class,Data,Input,Member,Output,Relationship,Design,Screen,MockUp,Function,Interface,State,Event,Epic,Feature,Goal,Idea,Issue,Req,Ticket,WorkPackage,Breakpoint,Barrier,Quality,Target,Scenario,Task,Test,Story,UseCase,VariationPoint,Variant},
+  emphstyle=\bfseries\color{entityColor},
+  emph={[2]has,is,superOf,binds,deprecates,excludes,helps,hurts,impacts,implements,interactsWith,precedes,requires,relatesTo,verifies},
+  emphstyle={[2]\color{relationColor}},
+  emph={[3]Attr,Code,Constraints,Comment,Deprecated,Example,Expectation,FileName,Gist,Image,Spec,Text,Title,Why,Benefit,Capacity,Cost,Damage,Frequency,Min,Max,Order,Prio,Probability,Profit,Value,Status},
+  emphstyle={[3]\itshape \color{attributeColor}},  
+}
+\lstset{style=reqT}
+
 \newcommand{\keywords}[1]{\par\addvspace\baselineskip
 \noindent\keywordname\enspace\ignorespaces#1}
-
 \begin{document}
@@
 \begin{abstract}
-[{\bf Context and motivation}] This research preview presents ongoing work on the metamodel of a free software requirements modeling tool called reqT that is developed in an educational context.
-[{\bf Question/problem}] The work aims to make an initial validation of a survey instrument that elicits views on the metamodel of the reqT tool, which seek to engage computer science students in Requirements Engineering (RE) through an open source requirements engineering DSL embedded in the Scala programming language.
-The research question underpinning the presented work is: Which RE concepts are essential to include in the metamodel for a requirements engineering tool in an educational context?
+[{\bf Context \& motivation}] This research preview presents ongoing work on the metamodel of a free software requirements modeling tool called reqT that is developed in an educational context.
+The work aims to make an initial validation of a survey instrument that elicits views on the metamodel of the reqT tool, which seek to engage computer science students in Requirements Engineering (RE) through an open source requirements engineering DSL embedded in the Scala programming language.
+[{\bf Question}] The research question is: Which RE concepts are essential to include in the metamodel for a requirements engineering tool in an educational context?
 [{\bf Principal ideas}] A survey instrument is developed with a list of 92 concepts (49 entities, 15 relations and 28 attributes) and a set of questions for each concept that elicit the respondents' views on the usage and interpretation of each concept.
@@
 The analysis of the responses suggest that many of the concepts in the metamodel are used frequently by the respondents and there is a large degree of agreement among the respondents about the meaning of the concepts.
+Some terms can be viewed as ''essential RE concepts'' in that a many use them and agree on their meaning.
 The results are encouraging for future work on empirical validation of the relevance of the reqT metamodel.
@@
 \section{Introduction}
-There are many challenges in teaching Requirements Engineering (RE) \cite{Memon2010, Regev2011}, including  conveying requirements modeling skills \cite{Callele2006}.
-Given a wide-spread attention on agile methods with less emphasis on extra-code artifacts \cite{Ramesh2010}, it may be particularly challenging to motivate coding-focused engineering students (and software practitioners) to spend serious effort on requirements modeling.
-One way to inspire software engineers to learn more about and do more RE may be to offer an interesting software tool.
+There are many challenges in teaching Requirements Engineering (RE) \cite{Memon2010, Regev2011}, including  conveying requirements modelling skills that can be used effectively in an unstructured, non-ideal, real-world situation \cite{Callele2006}.
+When teaching RE modelling we may ask ourselves: What are the \textit{essential} RE concepts that we should include in our taught metamodel for requirements?
+This paper investigates this questions in conjunction with the on-going work of developing a metamodel for reqT \cite{reqT}, an open source requirements engineering tool \cite{Regnell2013} used in RE education \cite{ets170}.
+A survey instrument is presented aiming to elicit the frequency of RE term usage and the degree of interpretation agreement.
+The responses from 15 Swedish RE scholars are analysed and discussed and conclusions suggest that a large subset of the concepts of the current reqT metamodel can be argued to be ''essential'' in that a majority of the subjects use them while agreeing with the concept definitions.
+The presented work is an initial validation and further work involving more subjects is needed to provide conclusions with more certainty.
+
+\section{Background and Related Work}
+
 There are nowadays numerous commercial RE tools available, but many are expensive, complex and not sufficiently open  \cite{Carillo2011}.
+A major aim of the reqT open source project is to provide a small but scalable, semi-formal and free software package for an educational setting \cite{Regnell2013} that can inspire code-loving computer science students to learn more about requirements modeling.
+The tool development started in 2011 at Lund University, where reqT is used in RE teaching at MSc level in the Computer Science \& Engineering program \cite{ets170}.
 
-This paper presents on-going work on a tool named reqT that aims to provide a small but scalable, semi-formal and free software package for an educational setting that (hopefully) can inspire code lovers to learn more about requirements modeling.
-A long-term goal of reqT is to offer an open platform for RE research prototypes, e.g.
-for feature modeling and release planning research.
-The tool development started in 2011 at Lund University, where reqT is used in RE teaching at MSc level in the Computer Science \& Engineering program.
-In 2012 reqT was rebuilt from scratch based on student feedback.
-The tool can be downloaded from: \url{http://reqT.org}
+A critical issue is how to find the "essential" RE concepts that allows for sufficient expressiveness, while not overloading the metamodel with esoteric concepts just for the sake of completeness. 
+%The reqT tool is used in a course based on a specific text book \cite{Lauesen2002} and a specific student project concept \cite{ets170}, and the concepts of the earlier versions of the reqT requirements metamodel reflect that context.
 
-The paper is organized as follows.
-Section \ref{section:objectives} states the objectives and motivates the design strategy of reqT.
-Section \ref{section:metamodel} presents the metamodel of reqT and some example reqT models.
-Section \ref{section:discussion} discusses limitations and some initial experiences from using reqT in teaching and concludes the paper with a sketch of future research directions.
+The reqT metamodel includes three types of concepts: entities, attributes and relations.
+Entities and attributes are nodes in a graph data structure, while relations are edges that can connect entities with sub-graphs.
+Thus a tree-like structure can be created of arbitrary depth spanning the graph that models some chunk of requirements.
 
-\section{Goals, Design Strategy and Rationale}\label{section:objectives}
-The main objective behind reqT is to establish a set of essential RE concepts and capture them in an expressive, extensible and executable language appealing to computer science students (and eventually practitioners).
-This general objective is accompanied by the following main goals and associated design strategies:
-\begin{enumerate}
-\item {\bf Semi-formal.} {\it Goal:} Provide a semi-formal representation of typical requirements modeling constructs that can illustrate a flexible combination of expressive natural language-style requirements with type-safe formalisms allowing static checks.
-{\it Design:} Use graph structures based on typed nodes representing typical requirement entities and attributes, and typed edges representing typical requirements relations, and implement the graph as an associative array (map).
-{\it Why?} Graphs are well-known to many CS students.
-Maps are efficient from an implementation perspective and may be less complex to master compared to e.g.
-SQL databases.
-\item {\bf Open.} {\it Goal:} Provide a platform-independent requirements tool that is free of charge.
-{\it Design:} Use Java Virtual Machine technology and release the code under an open source license.
-Use tab-separated, tabular text-files for import and export.
-Use HTML for document generation.
-{\it Why?} There are many free libraries available that runs on a JVM.
-Tab-sep and HTML support interoperability.
-%A free, platform-independent tool is appreciated by many students.
-\item {\bf Scalable.} {\it Goal:}  Provide an extensible requirements modeling language that can scale from small, concise models to large families of models with thousands of requirements entities and relations.
-{\it Design:} Implement reqT as an internal DSL (Domain-Specific Language) in the Scala programming language \cite{Odersky2004}.
-Use Map and Set from Scala collections to represent requirements graphs.
-{\it Why?} Scala is a modern, statically typed language with an efficient collections library.
-Scala offers scripting abilities that provide general extensibility without re-compilation.
-Integrated development environments \cite{eclipse}, as well as interactive scripting tools are available \cite{kojo}.
-%When requirements are expressed as code, general software engineering tools can be applied when scaling to large projects, e.g.
-configuration management tools and cloud-based code repositories to facilitate requirements evolution and collaboration over Internet.
-\end{enumerate}
+Version 3.0 of the metamodel includes the concepts defined in Appendix A.
+These concepts and definitions are gatherd from various sources including the IREB Glossary \cite{IREB}, wikipedia, terminology from agile development, variability modelling terminology, and the Lauesen text book \cite{Lauesen2002} used in the RE course at Lund University \cite{ets170}, in which reqT is applied in student role-playing projects.
 
-These goals, design strategies and rationale are directing the on-going work, and it remains to be investigated to what extent the main objective and goals can be met.
-A critical issue is how to interpret what are "essential" RE concepts and "typical" modeling constructs.
-The reqT tool is used in a course based on a specific text book \cite{Lauesen2002} and a specific student project concept \cite{ets170}, and the concepts of the reqT requirements metamodel (see Fig.
-\ref{fig:types}) reflect that context.
-However, the reqT architecture is prepared for extensions of new concepts in the metamodel to cater for different educational contexts.
-%%%%%%% THE MODEL
-\section{Modeling requirements with reqT}\label{section:metamodel}
-
-A reqT model includes sequences of graph parts {\tt<Entity><Edge><NodeSet>}  separated by comma and wrapped inside a {\tt Model( )} construct.
-A small reqT Model with three {\tt Feature} entities and one {\tt Stakeholder} entity is shown below:
+!!! EXPLAIN EXAMPLE !!!
 \begin{spacing}{0.8}
@@
     \fontsize{8pt}{12pt}\selectfont
-\begin{verbatim}
+\begin{lstlisting}
 Model(
-  Feature("f1") has (Spec("A good spec."), Status(SPECIFIED)),
-  Feature("f1") requires (Feature("f2"), Feature("f3")),
-  Stakeholder("s1") assigns(Prio(1)) to Feature("f2")
-)
-\end{verbatim}
+  Component("apperance") has (
+    VariationPoint("color") has (
+      Min(0), Max(2),
+      Variant("blue"), Variant("red"), Variant("green")),
+    VariationPoint("shape") has (
+      Min(1), Max(1), Variant("round"), Variant("square")),
+    VariationPoint("payment") has (
+      Min(1), Max(2), Variant("cash"), Variant("credit")),
+    VariationPoint("payment") requires Variant("cash"), 
+    Variant("round") excludes Variant("red"),
+    Variant("green") requires Variant("square")),
+  Component("apperance") requires VariationPoint("shape"), 
+  App("free") requires Component("apperance"),  
+  App("free") binds (
+    VariationPoint("shape") binds Variant("round")),
+  App("premium") requires Component("apperance"),  
+  App("premium") binds ( 
+    VariationPoint("color") binds (Variant("red"), Variant("green")),
+    VariationPoint("shape") binds (Variant("round"), Variant("square")),
+    VariationPoint("payment") binds Variant("cash")))
+\end{lstlisting}
 \endgroup
@@
 
+\section{Methodology and Data Collection}
+!!! explain how sources such as the IREB glossary was used to form the metamodel concept definitions !!!
+\section{Data Analysis}
+
+\textbf{Subject background.} The background questions in the survey regards the role of the subject, as shown in Table \ref{table:background}.
+In summary, the included\footnote{One subject answered NO on all background questions and was therefore excluded.} \input{../analysis/summary.tex}
+\begin{table}[H]
+\centering
+\fontsize{8}{9}\selectfont
+\caption{Background of subjects, $N = 15$. The table shows anonymized subject ids}
+\label{table:background}
+\begin{tabular}{p{0.42\textwidth}| p{0.58\textwidth}}
+\textit{Question} & \textit{Subject responding YES}  \\ \hline
+\input{../analysis/background.tex}
+
+ \end{tabular}
+\end{table}
+
+\textbf{Frequency analysis.} The idea of ''essentiality'' is characterized as the number of subjects that has responded that they (1) use the concepts at least in an informal, non-persistent way, and (2) use the concepts in a similar meaning as in the definition in Appendix A.
+Table \ref{table:frequency} shows the results of the frequency counts.
+
+%
+\begingroup
+\setlength{\tabcolsep}{4pt} % Default value: 6pt
+\renewcommand{\arraystretch}{1.5} % Default value: 1
+\begin{table}[H]
+\centering
+\fontsize{7}{8}\selectfont
+\caption{Frequency analysis of concepts' ''essentiality'', where $n$ is the number of subjects that answered $(use >= 1)$ and $(agree = 2)$ for each concept.}
+\label{table:frequency}
+\begin{tabular}{l | p{0.33\textwidth} | p{0.30\textwidth} | p{0.25\textwidth}}
+\textit{$n$} & \textit{Entities} & \textit{Attributes} & \textit{Relations} \\ \hline
+\input{../analysis/essential.tex}
+ \end{tabular}
+\end{table}
+\endgroup
 %%%%%%%%%%%%%% DISCUSSION
@@
 
- The results of the on-going work with reqT remains to be further investigated and a validation of reqT as a RE learning tool and research experimentation platform is subject to future work.
-This section discusses some preliminary experiences, limitations, relation to state-of-the-art and future research directions.
-
-{\bf Preliminary proof-of-concept.}
-The first version of reqT was tried on a voluntary basis by 12 students working in groups of 6 students each during fall 2011.
-Statements from course evaluations indicate that the students found reqT useful in their learning.
-One group used a configuration management tool for reqT models to manage  their parallel work, while one group used a cloud service and tab-sep export/import to collaborate over the Internet.
-The group with the largest requirements model produced 64 features, 18 tasks, 12 functions, 30 data requirements and 33 quality requirements, in total 157 requirements entities.
-
-Several students appreciated that reqT can mix informal text with a graph-oriented formalism, but some requested more elaborated functionality for document generation, as well as linking to external images.
-Some students also requested more modeling examples that show how the text book techniques could be transferred to reqT models.
-
-Based on student feedback, reqT was rebuilt from scratch during 2012 with a new architecture and a new version of the meta model (see Fig.
-\ref{fig:types}), as well as a revised Scala-internal DSL.
-The template-controlled HTML generation was implemented based on student suggestions.
-The teaching material was complemented with more example models directly related to the textbook.
-The second version of reqT is currently tested by  students in a new course instance and a post-course evaluation of reqT is planned in spring 2013.
-
-Our preliminary experiences from applying reqT in teaching suggest that reqT, if used in a suitable teaching context, may encourage students with a code-focused mind set to learn and practice RE in the following ways: (1) A free and platform-independent software tool that is implemented using a modern programming language with interactive scripting facilities can attracts the interest of code-focused students.
-(2) Requirements can be processed, queried, transformed or exported using Scala scripts, and the open-ended nature of reqT that allows students to code their own scripts to both manage requirements models and to adapt reqT to fit their RE needs in the course project was appreciated by several coding-literate students.
-(3) By turning requirements models into executable code, students can use programming tools such as a console command line interpreter (the Scala REPL) as well as a source code version control system (e.g.
-git-scm.com) to branch and merge their collaborative work on requirements in ways they are used to from their previous collaborative software implementation courses, including issue tracking systems and code review support.
-
-{\bf Relation to state-of-the-art.} To the best of our knowledge there is no other RE tool that allows semi-formal requirement models to become executable programs through an internal Scala DSL, and thus letting coding, testing and requirements engineering share the same media.
-In the RE tool survey by Carrillo de Gea et al.
-\cite{Carillo2011} it is pointed out that "many expensive tools aren't sufficiently open".
-The reqT technology aims to be completely free and open to facilitate academic usage, collaborative evolution and incorporation of new RE concepts in different teaching and research contexts.
-Many of the existing tools have proprietary representations \cite{Carillo2011}, while users of reqT can extend the reqT metamodel with new entities and attributes simply by adding case classes with a few lines of code.
-However, reqT cannot compete with versatile commercial RE tools  \cite{Carillo2011}  in terms of e.g.
-features completeness and graphical user interface.
+!!Discuss and conclude!!
 
 {\bf Limitations.
-} In its current version, reqT has a number of limitations: (1) As the user interface is text based and depends on the command line interface of the Scala REPL or a script editor environment \cite{kojo, eclipse}, students that only are prepared to use graphical user interfaces may be discouraged.
-Some of our students preferred to work in a GUI spreadsheet application using tab-separated exports from reqT that was generated by other team members assigned by the student group to be reqT experts.
-(2) It requires some knowledge of Scala to tailor reqT exports and there is a need for a more comprehensive API for adaptable document generation.
-(3) The embedded DSL requires some learning efforts and it remains to be investigated if the effort is justified by the knowledge gained.
-(4) To support scalability to large families of reqT models there is a need for modularization concepts and overview visualizations.
-(5) The explicit typing of entities with keywords such Feature and Stakeholder can be perceived as verbose compared to more concise but potentially cryptic abbreviations (e.g.
-Fe, Sh).
-This may be addressed by DSL-specific editor support, such as code-completion, code folding and code templates.
+} Due to the limited number of subjects and the high degree of homogeneity among subjects with respect to background, it is difficult to analyse and draw conclusions about potential differences in opinions between e.g.
+teachers and developers.
+Some subjects needed more time and completed survey offline during the coming days, which may give a variation in how careful the responses were considered.
+The survey was conducted by the author and inventor of reqT, in conjunction with a seminar and demo of reqT.
+In order to avoid any positive bias due to advocacy in favour of the reqT metamodel, the survey was held prior to the seminar and demo.
+This in turn may introduce a threat of limited knowledge among subjects of the idea behind the modelling approach in reqT
 
-{\bf Future work.} Further directions of research include (1) incorporation of constraints on models for support of prioritization and release planning \cite{Regnell2011}, (2) more elaborate semantic checks to better guide requirements modelers, and (3) graphical visualization of requirements graph models.
+{\bf Future work.} %Further directions of research include (1) incorporation of constraints on models for support of prioritization and release planning \cite{Regnell2011}, (2) more elaborate semantic checks to better guide requirements modelers, and (3) graphical visualization of requirements graph models.
 (4) Natural Language Processing technology including e.g.
@@
 
-A major objective of this research preview paper is to expose the latest version of reqT to the community of RE scholars and to invite discussions and contributions.
+%A major objective of this research preview paper is to expose the latest version of reqT to the community of RE scholars and to invite discussions and contributions.
+
+\subsubsection*{Acknowledgments.} This work is partly funded by  VINNOVA within the EASE project.
 
@@
 
+%%%%%%%%%%%%%%% APPENDIX WITH DEFINITIONS
 \clearpage
-\hspace{-0.5cm}
+%\hspace{-0.9cm}
+\section*{Appendix A: Definitions of Metamodel Concepts}
 \begin{minipage}{1\linewidth}
-\begin{minipage}{0.36\textwidth} 
+\begin{minipage}{0.30\textwidth}
+
 \begin{table}[H]
 %\centering
-\vspace{-0.7cm}
+\vspace{-0.8cm}
 \fontsize{6}{6.5}\selectfont
-\caption{Definitions of concepts.}%: Entities, Attributes and Relations.}
-\label{table-entities}
-\begin{tabular}{|l p{5.0cm}|}
+%\caption{.}%: Entities, Attributes and Relations.}
+
+\begin{tabular}{|l p{4.5cm}|}
 \hline
 \textit{Entity} & \textit{Definition}  \\ \hline
-Actor&A human or machine that communicates with a system.\\
-App&A computer program, or group of programs designed for end users, normally with a graphical user interface. Short for application.\\
-Barrier&Something that makes it difficult to achieve a goal or a higher quality level.\\
-Breakpoint&A point of change. An important aspect of a (non-linear) relation between quality and benefit.\\
-Class&An extensible template for creating objects. A set of objects with certain attributes in common. A category.\\
-Component&A composable part of a system. A reusable, interchangeable system unit or functionality.\\
-Configuration&A specific combination of variants.\\
-Data&Information stored in a system.\\
-Design&A specific realization or high-level implementation description (of a system part).\\
-Domain&The application area of a product with its surrounding entities.\\
-Epic&A large user story or a collection of stories.\\
-Event&Something that can happen in the domain and/or in the system.\\
-Feature&A releasable characteristic of a product. A (high-level, coherent) bundle of requirements.\\
-Function&A description of how input data is mapped to output data. A capability of a system to do something specific.\\
-Goal&An intention of a stakeholder or desired system property.\\
-Idea&A concept or thought (potentially interesting).\\
-Interface&A defined way to interact with a system.\\
-Issue&Something needed to be fixed.\\
-Item& An article in a collection, enumeration, or series.\\
-Label&A descriptive name used to identify something.\\
-Member&An entity that is part of another entity, eg. a field in a in a class.\\
-Meta&A prefix used on a concept to mean beyond or about its own concept, e.g. metadata is data about data.\\
-MockUp&A prototype with limited functionality used to demonstrate a design idea.\\
-Module&A collection of coherent functions and interfaces.\\
-Product&Something offered to a market.\\
-Quality&A distinguishing characteristic or degree of goodness.\\
-Relationship&A specific way that entities are connected.\\
-Release&A specific version of a system offered at a specific time to end users.\\
-Req&Something needed or wanted. An abstract term denoting any type of information relevant to the (specification of) intentions behind system development. Short for requirement.\\
-Resource&A capability of, or support for development.\\
-Risk&Something negative that may happen.\\
-Scenario&A (vivid) description of a (possible future) system usage.\\
-Screen&A design of (a part of) a user interface.\\
-Section&A part of a (requirements) document.\\
-Service&Actions performed by systems and/or humans to provide results to stakeholders.\\
-Stakeholder&Someone with a stake in the system development or usage.\\
-State&A mode or condition of something in the domain and/or in the system. A configuration of data.\\
-Story&A short description of what a user does or needs. Short for user story.\\
-System&A set of interacting software and/or hardware components.\\
-Target&A desired quality level or goal .\\
-Task&A piece of work (that users do, maybe supported by a system).\\
-Term&A word or group of words having a particular meaning.\\
-Test&A procedure to check if requirements are met.\\
-Ticket&(Development) work awaiting to be completed.\\
-UseCase&A list of steps defining interactions between actors and a system to achieve a goal.\\
-User&A human interacting with a system.\\
-Variant&An object or system property that can be chosen from a set of options.\\
-VariationPoint&An opportunity of choice among variants.\\
-WorkPackage&A collection of (development) work tasks.\\
+\input{../analysis/entity-defs.tex}
 \hline
@@
 %\caption{Definitions of attributes and relations.}
-\vspace{-0.2cm}
+\vspace{-0.8cm}
 \label{table-entities}
-\begin{tabular}{|l p{4.3cm}|}
+\begin{tabular}{|l p{4.0cm}|}
  \hline
 \textit{Attribute} & \textit{Definition}  \\ \hline
-Benefit&A characterisation of a good or helpful result or effect (e.g. of a feature).\\
-Capacity&The largest amount that can be held or contained (e.g. by a resource).\\
-Code&A collection of (textual) computer instructions in some programming language, e.g. Scala. Short for source code.\\
-Comment&A note that explains or discusses some entity.\\
-Constraints&A collection of propositions that restrict the possible values of a set of variables.\\
-Cost&The expenditure of something, such as time or effort, necessary for the implementation of an entity.\\
-Damage&A characterisation of the negative consequences if some entity (e.g. a risk) occurs.\\
-Deprecated&A description of why an entity should be avoided, often because it is superseded by another entity, as indicated by a 'deprecates' relation.\\
-Example&A note that illustrates some entity by a  typical instance.\\
-Expectation&The required output of a test in order to be counted as passed.\\
-FileName&The name of a storage of serialized, persistent data.\\
-Frequency&The rate of occurrence of some entity. \\
-Gist&A short and simple description of an entity, e.g. a function or a test.\\
-Image&(The name of) a picture of an entity.\\
-Input&Data consumed by an entity, \\
-Max&The maximum estimated or assigned (relative) value.\\
-Min&The minimum estimated or assigned (relative) value.\\
-Order&The ordinal number of an entity (1st, 2nd, ...).\\
-Output&Data produced by an entity, e.g. a function or a test.\\
-Prio&The level of importance of an entity. Short for priority.\\
-Probability&The likelihood that something (e.g. a risk) occurs.\\
-Profit&The gain or return of some entity, e.g. in monetary terms.\\
-Spec&A (detailed) definition of an entity. Short for specification\\
-Status&A level of refinement of an entity (e.g. a feature) in the development process. \\
-Text&A sequence of words (in natural language).\\
-Title&A general or descriptive heading.\\
-Value&An amount. An estimate of worth.\\
-Why&A description of intention. Rationale.\\
+\input{../analysis/attribute-defs.tex}
 
@@
 \textit{Relation} & \textit{Definition}  \\ \hline
-binds&Ties a value to an option. A configuration binds a variation point.\\
-deprecates&Makes outdated. An entity deprecates (supersedes) another entity.\\
-excludes&Prevents a combination. An entity excludes another entity.\\
-has&Expresses containment, substructure. An entity contains another entity.\\
-helps&Positive influence. A goal helps to fulfil another goal.\\
-hurts&Negative influence. A goal hinders another goal.\\
-impacts&Some influence. A new feature impacts an existing component.\\
-implements&Realisation of. A module implements a feature.\\
-interactsWith&Communication. A user interacts with an interface.\\
-is&Sub-typing, specialization, part of another, more general entity.\\
-precedes&Temporal ordering. A feature precedes (is implemented before) another feature.\\
-relatesTo&General relation. An entity is related to another entity.\\
-requires&Requested combination. An entity is required (or wished) by another entity.\\
-superOf&Super-typing, generalization, includes another, more specific entity.\\
-verifies&Gives evidence of correctness. A test verifies the implementation of a feature.\\
+\input{../analysis/relation-defs.tex}
  \hline
@@
 \end{table}
+
 \end{minipage}
 \end{minipage}
-\subsubsection*{Acknowledgments.} This work is partly funded by  VINNOVA within the EASE project.
 

### 2015-10-01 src/script.scala

@@ -21,7 +21,7 @@ case class Matrix[T](vectors: Vector[Vector[T]]){
   def get(x: Int, y: Int): Option[T] = Try(vectors(x)(y)).toOption
 }
 
-object Xls {
+object Xls {  // ----------- Load xls file into a Matrix ---------------------
 
   def load(fileName: String, sheetNum: Int = 0): Matrix[String] = 
     loadFile(new File(fileName), sheetNum)
@@ -45,8 +45,17 @@ object Xls {
     if (cell != null) dataFormatter.formatCellValue(cell) else ""
 }
 
-object Extract { 
-  val DATA_DIR = "Data"
+object Extract {  // ----------- Process the Matrix and output results -----------------
+  val DATA_DIR = "data/"
+  val OUT_DIR = "analysis/"
+  val FILE_SUMMARY = OUT_DIR + "summary.tex"
+  val TABLE_ENT = OUT_DIR + "entity-defs.tex"
+  val TABLE_ATTR = OUT_DIR + "attribute-defs.tex"
+  val TABLE_REL = OUT_DIR + "relation-defs.tex"
+  val FILE_BACKGROUND = OUT_DIR + "background.tex"
+  val FILE_ESSENTIAL = OUT_DIR + "essential.tex"
+  val NL = "\\\\"
+
   def toInt(s: String): Int = Try(s.toDouble.toInt).getOrElse(0) 
   case class Quest(usageStr: String, meaningStr: String, other: String, synonym: String){
     val usage = toInt(usageStr)
@@ -54,6 +63,9 @@ object Extract {
     val agreement = usage + meaning
   }
   def ls(dir: String) = (new File(dir)).listFiles.toVector
+  def resetFiles(files: String*) = files.map(scala.tools.nsc.io.File(_).writeAll(""))
+  def writelnToFile(file: String)(data: String) = scala.tools.nsc.io.File(file).writeAll(data+"\n")
+  def appendlnToFile(file: String)(data: String) = scala.tools.nsc.io.File(file).appendAll(data+"\n")
   val files = ls(DATA_DIR)
   val data = files.map(Xls.loadFile(_))
 
@@ -68,11 +80,12 @@ object Extract {
   val (teach, develop, research) = (9, 10, 11)
   def yesOf(xs: Vector[String]) = xs.zipWithIndex.filter(_._1.toLowerCase == "yes").map(_._2)
[…]

### 2015-10-01 paper/paper-refsq2016-submission.tex
add stuff to paper
add more stuff to the paper

@@
 % first the title is needed
-\title{What is essential? -- A pilot survey on views about the requirements metamodel of reqT}
+\title{What is essential? -- A pilot survey on views about the requirements metamodel of reqT.org}
 %reqT --  A Requirements Modeling Tool for Code Lovers
@@
 
-\keywords{requirements engineering, requirements metamodel, CASE tool, requirements engineering education, embedded domain-specific language, empirical software engineering}
+\keywords{requirements engineering, metamodel, CASE tool, engineering education, embedded domain-specific language, empirical software engineering.}
 \end{abstract}
@@
 A survey instrument is presented aiming to elicit the frequency of RE term usage and the degree of interpretation agreement.
-The responses from 15 Swedish RE scholars are analysed and discussed and conclusions suggest that a large subset of the concepts of the current reqT metamodel can be argued to be ''essential'' in that a majority of the subjects use them while agreeing with the concept definitions.
+The responses from 14 Swedish RE scholars are analysed and discussed and conclusions suggest that a large subset of the concepts of the current reqT metamodel can be argued to be ''essential'' in that a majority of the subjects use them while agreeing with the concept definitions.
 The presented work is an initial validation and further work involving more subjects is needed to provide conclusions with more certainty.
@@
 
-A critical issue is how to find the "essential" RE concepts that allows for sufficient expressiveness, while not overloading the metamodel with esoteric concepts just for the sake of completeness. 
+A critical issue is how to choose the "essential" RE concepts that allows for sufficient expressiveness, while not overloading the metamodel with esoteric concepts just for the sake of completeness. 
 %The reqT tool is used in a course based on a specific text book \cite{Lauesen2002} and a specific student project concept \cite{ets170}, and the concepts of the earlier versions of the reqT requirements metamodel reflect that context.
@@
 
-Version 3.0 of the metamodel includes the concepts defined in Appendix A.
-These concepts and definitions are gatherd from various sources including the IREB Glossary \cite{IREB}, wikipedia, terminology from agile development, variability modelling terminology, and the Lauesen text book \cite{Lauesen2002} used in the RE course at Lund University \cite{ets170}, in which reqT is applied in student role-playing projects.
+The code below shows a toy example of an orthogonal variability model \cite{metzger2007variability} expressed in the reqT Scala-embedded DSL \cite{Regnell2013} illustrating a small part of its metamodel.
+Other parts of the metamodel contains concepts that enable e.g.
+goal modelling %\cite{liu2004}
+and use case and user story modelling, see further Appendix A.
 
-!!! EXPLAIN EXAMPLE !!!
 \begin{spacing}{0.8}
@@
 \end{spacing}
+ Entities are in bold, attributes in italics and relations start with a lower case letter.
+In the reqT editor, entities, attributes, and relations are syntax-coloured in blue, green and red respectively.
+A reqT model written in the above syntax is actually valid Scala code that, when executed, generates an immutable data structure that can be traversed and manipulated using Scala scripts.
+Also visualisations can be generated using GraphViz export and export to HTML and spreadsheet formats.
 
 \section{Methodology and Data Collection}
-!!! explain how sources such as the IREB glossary was used to form the metamodel concept definitions !!!
+In order to validate RE scholar's opinions of the metamodel a survey instrument is developed including 49 entities, 15 relations and 28 attributes of the reqT Version 3.0 metamodel.
+All the concepts are listed with their definitions in Appendix A.\footnote{The survey is available at \url{https://github.com/reqT/reqT/tree/3.0.x/survey}} These concepts and definitions were gathered from various sources including the IREB Glossary \footnote{\url{https://www.ireb.org/en/cpre/cpre-glossary/}}, wikipedia, terminology from agile development, variability \cite{metzger2007variability} and goal modelling terminology, and the Lauesen text book \cite{Lauesen2002} used in the RE course at Lund University \cite{ets170}, in which reqT is applied in student role-playing projects.
+\begin{figure}[h]
+\centering
+\includegraphics[width=\textwidth]{img/survey-screen-dump}
+\caption{A screen dump of a part of the survey instrument.}
+\label{fig:survey}
+\end{figure}
+
+\noindent The data collection for the presented pilot run was made during a Swedish national network meeting with academic RE scholars in the spring of 2015.
+The survey was filled in during the meeting using the participants' own laptops  in a downloadable spreadsheet as shown in Figure~\ref{fig:survey}.
+The subjects were given 20 minutes to complete the survey.
+Most of the subjects handed in the survey via email directly after the session, while a few finshed it offline and emailed their response after the meeting.
 \section{Data Analysis}
@@
 \fontsize{8}{9}\selectfont
-\caption{Background of subjects, $N = 15$. The table shows anonymized subject ids}
+\caption{Background of subjects, $N = 15$. The table use the subjects' anonymous ids S01--S15.}
 \label{table:background}
-\begin{tabular}{p{0.42\textwidth}| p{0.58\textwidth}}
-\textit{Question} & \textit{Subject responding YES}  \\ \hline
+\begin{tabular}{p{0.43\textwidth}| p{0.58\textwidth}}
+\textit{Background question} & \textit{Subject responding YES}  \\ 
 \input{../analysis/background.tex}
@@
 
-\textbf{Frequency analysis.} The idea of ''essentiality'' is characterized as the number of subjects that has responded that they (1) use the concepts at least in an informal, non-persistent way, and (2) use the concepts in a similar meaning as in the definition in Appendix A.
+\noindent \textbf{Frequency analysis.} The idea of ''essentiality'' is characterized as the number of subjects that has responded that they (1) use the concepts at least in an informal or non-persistent way, \textit{and} that they (2) use the concepts in a similar meaning as in the definition in Appendix A.
+Figure~\ref{fig:survey} shows the definitions of the three-level ordinal scales of Questions $Q1_{usage}$  and $Q2_{meaning}$ respectively.
 Table \ref{table:frequency} shows the results of the frequency counts.
+If an ''essentiality threshold'' is chosen at $N/2$ then only the $9$ concepts from row $n = 7$ and below in Table~\ref{table:frequency} are considered ''non-essential'', hence showing that more than 90\% of the metamodel concepts have a majority of the subjects that use and agree upon them.
+All concepts have at least one subject that use it and agree with the definition.
 
@@
 \setlength{\tabcolsep}{4pt} % Default value: 6pt
-\renewcommand{\arraystretch}{1.5} % Default value: 1
-\begin{table}[H]
+\renewcommand{\arraystretch}{1.2} % Default value: 1
+\begin{table}
 \centering
 \fontsize{7}{8}\selectfont
-\caption{Frequency analysis of concepts' ''essentiality'', where $n$ is the number of subjects that answered $(use >= 1)$ and $(agree = 2)$ for each concept.}
+\caption{Frequency analysis, where $n$ is the number of subjects that for the respective concepts answered $(Q1_{usage}>= 1)$ \textit{and} $(Q2_{meaning} = 2)$.
+In total there are 92 concepts (49 entities, 15 relations and 28 attributes).
+The higher up in the table, the more ''essential''.}
 \label{table:frequency}
@@
 \endgroup
+
 %%%%%%%%%%%%%% DISCUSSION
@@
 
-!!Discuss and conclude!!
+It can be questioned if ''essentiallity'' of a set of RE concepts can be characterized by how many RE scholars that use them.
+However, if someone use a certain concept and wants to model it, then the metamodel of the applied modelling approach needs to have it, in order for that person not to find that the metamodel is lacking vital parts.
+The presented survey is a pilot investigation with two main contributions: (1) the survey instrument and the data collection and analysis approach is feasible, and (2) for more than 90\% of the metamodel concept there is a majority of the 14 participating RE scholars that use them and agree upon their definition.
 
@@
 teachers and developers.
-Some subjects needed more time and completed survey offline during the coming days, which may give a variation in how careful the responses were considered.
-The survey was conducted by the author and inventor of reqT, in conjunction with a seminar and demo of reqT.
+Some subjects needed more time and completed their survey offline during the coming days, which may give a variation in how careful the responses were considered.
+%The survey was conducted by the author and inventor of reqT, in conjunction with a seminar and demo of reqT.
 In order to avoid any positive bias due to advocacy in favour of the reqT metamodel, the survey was held prior to the seminar and demo.
@@
 
-{\bf Future work.} %Further directions of research include (1) incorporation of constraints on models for support of prioritization and release planning \cite{Regnell2011}, (2) more elaborate semantic checks to better guide requirements modelers, and (3) graphical visualization of requirements graph models.
+{\bf Further work.} When developing a metamodel it is interesting not just to ask if the concepts to include are essential, but also to pose the question if is the set of concepts is complete.
+If some essential concept is missing from some stakleholder's viewpoint, then the metamodel is not sufficient.
+With more subjects participating in the presented RE metamodel survey, the analysis of answers to further questions on alternative terms and missing concepts will be enabled and beneficial to the further development of a comprehensive and complete, but not overloaded, RE metamodel.
+%Further directions of research include (1) incorporation of constraints on models for support of prioritization and release planning \cite{Regnell2011}, (2) more elaborate semantic checks to better guide requirements modelers, and (3) graphical visualization of requirements graph models.
 (4) Natural Language Processing technology including e.g.
@@
 %\centering
-\vspace{-0.8cm}
+\vspace{-1.0cm}
 \fontsize{6}{6.5}\selectfont
@@
 
-\begin{tabular}{|l p{4.5cm}|}
-\hline
-\textit{Entity} & \textit{Definition}  \\ \hline
+\begin{tabular}{l p{4.6cm}}
+
+\textbf{\textit{Entity}} & \textbf{\textit{Definition}}  \\
 \input{../analysis/entity-defs.tex}
-\hline
  \end{tabular}
@@
  \hspace{4em plus 1fill} 
-\begin{minipage}{0.47\textwidth} 
+\begin{minipage}{0.46\textwidth} 
  \begin{table}[H]
@@
 %\caption{Definitions of attributes and relations.}
-\vspace{-0.8cm}
+\vspace{-1.0cm}
 \label{table-entities}
-\begin{tabular}{|l p{4.0cm}|}
- \hline
-\textit{Attribute} & \textit{Definition}  \\ \hline
+\begin{tabular}{|l p{4.1cm}}
+
+\textbf{\textit{Attribute}} & \textbf{\textit{Definition}}  \\ 
 \input{../analysis/attribute-defs.tex}
+ & \\
+ \textbf{\textit{Relation}} & \textbf{\textit{Definition}}  \\ 
+\input{../analysis/relation-defs.tex}
 
-\hline
-\textit{Relation} & \textit{Definition}  \\ \hline
-\input{../analysis/relation-defs.tex}
- \hline
 \end{tabular}

### 2015-10-01 src/script.scala

@@ -80,7 +80,7 @@ object Extract {  // ----------- Process the Matrix and output results ---------
   val (teach, develop, research) = (9, 10, 11)
   def yesOf(xs: Vector[String]) = xs.zipWithIndex.filter(_._1.toLowerCase == "yes").map(_._2)
   def background = for (i <- Seq(teach, develop, research)) yield {
-     val ids = yesOf(data.map(_.apply(i, 4))).map(_ + 1).map(_.toString).map(s => "R" + "0"*(2-s.length) + s)
+     val ids = yesOf(data.map(_.apply(i, 4))).map(_ + 1).map(_.toString).map(s => "S" + "0"*(2-s.length) + s)
      val question = data(0)(i,1) + " YES/NO"
      (question, ids)
   } 

### 2015-10-01 paper/paper-refsq2016-submission.tex
add more stuff to the paper

@@
 \end{spacing}
- Entities are in bold, attributes in italics and relations start with a lower case letter.
+\noindent Entities are in bold, attributes in italics and relations start with a lower case letter.
 In the reqT editor, entities, attributes, and relations are syntax-coloured in blue, green and red respectively.
-A reqT model written in the above syntax is actually valid Scala code that, when executed, generates an immutable data structure that can be traversed and manipulated using Scala scripts.
+A reqT model written in the above syntax is actually valid Scala code that, when executed, generates data structure that can be traversed and manipulated using Scala scripts.
 Also visualisations can be generated using GraphViz export and export to HTML and spreadsheet formats.

### 2015-10-01 paper/paper-refsq2016-submission.tex
add more stuff to the paper

@@
   %breakatwhitespace=true,
-  basicstyle=\ttfamily\scriptsize,
+  basicstyle=\ttfamily\fontsize{8}{10}\selectfont,
   emph={Ent,Meta,Item,Label,Section,Term,Actor,App,Component,Domain,Module,Product,Release,Resource,Risk,Service,Stakeholder,System,User,Class,Data,Input,Member,Output,Relationship,Design,Screen,MockUp,Function,Interface,State,Event,Epic,Feature,Goal,Idea,Issue,Req,Ticket,WorkPackage,Breakpoint,Barrier,Quality,Target,Scenario,Task,Test,Story,UseCase,VariationPoint,Variant},
@@
 [{\bf Contribution}] The survey is initially validated in a pilot study involving 14 Swedish RE scholars as subjects.
-The survey results indicate that the survey is feasible if the respondents is willing to invest around 30 minutes of their time.
+The survey results indicate that the survey is feasible.
 The analysis of the responses suggest that many of the concepts in the metamodel are used frequently by the respondents and there is a large degree of agreement among the respondents about the meaning of the concepts.
-Some terms can be viewed as ''essential RE concepts'' in that a many use them and agree on their meaning.
 The results are encouraging for future work on empirical validation of the relevance of the reqT metamodel.
@@
 
-\section{Background and Related Work}
+\section{Background}
 
@@
 
-\begin{spacing}{0.8}
-\begingroup
-    \fontsize{8pt}{12pt}\selectfont
+\begin{spacing}{1.1}
 \begin{lstlisting}
@@
 \end{lstlisting}
-\endgroup
 \end{spacing}
@@
 \setlength{\tabcolsep}{4pt} % Default value: 6pt
-\renewcommand{\arraystretch}{1.2} % Default value: 1
+\renewcommand{\arraystretch}{1.35} % Default value: 1
 \begin{table}

### 2015-10-02 paper/paper-refsq2016-submission.tex
fix spelling and wording before submission

@@
 \usepackage{hyperref}
+\usepackage{cleveref} % http://tex.stackexchange.com/questions/10102/multiple-references-to-the-same-footnote-with-hyperref-support-is-there-a-bett/10116#10116
+\crefformat{footnote}{#2\footnotemark[#1]#3}
+
 \usepackage{xcolor}
@@
 \begin{abstract}
-[{\bf Context \& motivation}] This research preview presents ongoing work on the metamodel of a free software requirements modeling tool called reqT that is developed in an educational context.
+[{\bf Context \& motivation}] This research preview paper presents ongoing work on the metamodel of a free software requirements modeling tool called reqT that is developed in an educational context.
 The work aims to make an initial validation of a survey instrument that elicits views on the metamodel of the reqT tool, which seek to engage computer science students in Requirements Engineering (RE) through an open source requirements engineering DSL embedded in the Scala programming language.
@@
 \section{Introduction}
-There are many challenges in teaching Requirements Engineering (RE) \cite{Memon2010, Regev2011}, including  conveying requirements modelling skills that can be used effectively in an unstructured, non-ideal, real-world situation \cite{Callele2006}.
-When teaching RE modelling we may ask ourselves: What are the \textit{essential} RE concepts that we should include in our taught metamodel for requirements?
-This paper investigates this questions in conjunction with the on-going work of developing a metamodel for reqT \cite{reqT}, an open source requirements engineering tool \cite{Regnell2013} used in RE education \cite{ets170}.
+There are many challenges in teaching Requirements Engineering (RE) \cite{Memon2010, Regev2011}, including  advancing students' requirements modelling skills that can be used effectively in an unstructured, non-ideal, real-world situation \cite{Callele2006}.
+When teaching RE modelling we may ask ourselves: What are the \textit{essential} RE concepts that we should include in a taught metamodel for requirements?
+This paper investigates this questions in conjunction with the on-going work of developing a metamodel for reqT \cite{reqT}, an open source requirements engineering tool used in RE education  \cite{Regnell2013}.
 A survey instrument is presented aiming to elicit the frequency of RE term usage and the degree of interpretation agreement.
-The responses from 14 Swedish RE scholars are analysed and discussed and conclusions suggest that a large subset of the concepts of the current reqT metamodel can be argued to be ''essential'' in that a majority of the subjects use them while agreeing with the concept definitions.
-The presented work is an initial validation and further work involving more subjects is needed to provide conclusions with more certainty.
+The responses from 14 Swedish RE scholars are analysed and discussed and conclusions suggest that a large subset of the concepts of the current reqT metamodel can be seen as ''essential'' in that a majority of the subjects use them while agreeing with the concepts' definitions.
+The presented work represents an initial validation of the survey instrument.
+Further work involving more subjects is needed to draw conclusions with more certainty.
 
@@
 A major aim of the reqT open source project is to provide a small but scalable, semi-formal and free software package for an educational setting \cite{Regnell2013} that can inspire code-loving computer science students to learn more about requirements modeling.
-The tool development started in 2011 at Lund University, where reqT is used in RE teaching at MSc level in the Computer Science \& Engineering program \cite{ets170}.
+The tool development started in 2011 at Lund University, where reqT is used in RE teaching at MSc level in student role-playing projects.\footnote{\label{note:ets170}The Lund Univ.
+MSc-level RE course can be found at: \scriptsize\url{http://cs.lth.se/education}}
 
-A critical issue is how to choose the "essential" RE concepts that allows for sufficient expressiveness, while not overloading the metamodel with esoteric concepts just for the sake of completeness. 
+A critical issue is how to choose the essential RE concepts that allows for sufficient expressiveness, while not overloading the metamodel with esoteric concepts just for the sake of completeness. 
 %The reqT tool is used in a course based on a specific text book \cite{Lauesen2002} and a specific student project concept \cite{ets170}, and the concepts of the earlier versions of the reqT requirements metamodel reflect that context.
@@
 Other parts of the metamodel contains concepts that enable e.g.
-goal modelling %\cite{liu2004}
-and use case and user story modelling, see further Appendix A.
+goal modelling, %\cite{liu2004}
+use case modelling, and user story modelling, see further Appendix A.
 
@@
     VariationPoint("color") has (
-      Min(0), Max(2),
-      Variant("blue"), Variant("red"), Variant("green")),
+      Min(0), Max(2), Variant("blue"), Variant("red"), Variant("green")),
     VariationPoint("shape") has (
@@
   Component("apperance") requires VariationPoint("shape"), 
-  App("free") requires Component("apperance"),  
-  App("free") binds (
-    VariationPoint("shape") binds Variant("round")),
-  App("premium") requires Component("apperance"),  
+  App("free") has Component("apperance"),  
+  App("free") binds (VariationPoint("shape") binds Variant("round")),
+  App("premium") has Component("apperance"),  
   App("premium") binds ( 
@@
 \end{spacing}
-\noindent Entities are in bold, attributes in italics and relations start with a lower case letter.
+\noindent Entities in the above code listing are in bold, attributes in italics and relations start with a lower case letter.
 In the reqT editor, entities, attributes, and relations are syntax-coloured in blue, green and red respectively.
-A reqT model written in the above syntax is actually valid Scala code that, when executed, generates data structure that can be traversed and manipulated using Scala scripts.
+A reqT model written in the above syntax is actually valid Scala code that, when executed, generates a data structure that can be traversed and manipulated using Scala scripts.
 Also visualisations can be generated using GraphViz export and export to HTML and spreadsheet formats.
@@
 \section{Methodology and Data Collection}
-In order to validate RE scholar's opinions of the metamodel a survey instrument is developed including 49 entities, 15 relations and 28 attributes of the reqT Version 3.0 metamodel.
-All the concepts are listed with their definitions in Appendix A.\footnote{The survey is available at \url{https://github.com/reqT/reqT/tree/3.0.x/survey}} These concepts and definitions were gathered from various sources including the IREB Glossary \footnote{\url{https://www.ireb.org/en/cpre/cpre-glossary/}}, wikipedia, terminology from agile development, variability \cite{metzger2007variability} and goal modelling terminology, and the Lauesen text book \cite{Lauesen2002} used in the RE course at Lund University \cite{ets170}, in which reqT is applied in student role-playing projects.
+In order to validate RE scholar's opinions of the metamodel, a survey instrument was developed including the 49 entities, 15 relations and 28 attributes of the reqT Version 3.0 metamodel.
+All the concepts are listed with their definitions in Appendix A.\footnote{The survey is available at \scriptsize\url{https://github.com/reqT/reqT/tree/3.0.x/survey}} These 92 concepts and definitions were gathered from various sources including the IREB Glossary \footnote{\scriptsize\url{https://www.ireb.org/en/cpre/cpre-glossary/}}, wikipedia, terminology from agile development, variability \cite{metzger2007variability} and goal modelling terminology, and the text book\cite{Lauesen2002} used in an RE course at Lund Univ.\cref{note:ets170}
+\vspace{-0.2cm}
 \begin{figure}[h]
@@
 \noindent The data collection for the presented pilot run was made during a Swedish national network meeting with academic RE scholars in the spring of 2015.
-The survey was filled in during the meeting using the participants' own laptops  in a downloadable spreadsheet as shown in Figure~\ref{fig:survey}.
-The subjects were given 20 minutes to complete the survey.
-Most of the subjects handed in the survey via email directly after the session, while a few finshed it offline and emailed their response after the meeting.
+The survey was filled in during the meeting using the participants' own laptops  in a downloadable spreadsheet as shown in Fig.~\ref{fig:survey}.
+The subjects were given around 20 minutes to complete the survey.
+Most of the subjects handed in the survey via email directly after the session, while a few finished it offline and emailed their responses after the meeting.
 \section{Data Analysis}
@@
 \textbf{Subject background.} The background questions in the survey regards the role of the subject, as shown in Table \ref{table:background}.
-In summary, the included\footnote{One subject answered NO on all background questions and was therefore excluded.} \input{../analysis/summary.tex}
+The analyzed\footnote{One subject answered NO on all background questions and was therefore excluded.} \input{../analysis/summary.tex} The response rate was 100\% after a reminder was emailed to one missing subject.
+
+\begingroup
 \begin{table}[H]
+\setlength{\tabcolsep}{4pt} % Default value: 6pt
+\renewcommand{\arraystretch}{1.4} % Default value: 1
 \centering
-\fontsize{8}{9}\selectfont
-\caption{Background of subjects, $N = 15$. The table use the subjects' anonymous ids S01--S15.}
+\fontsize{8.5}{9}\selectfont
+\caption{Background of subjects, $N = 15$. The subjects were given anonymous ids S01--S15.}
 \label{table:background}
-\begin{tabular}{p{0.43\textwidth}| p{0.58\textwidth}}
-\textit{Background question} & \textit{Subject responding YES}  \\ 
+\begin{tabular}{p{0.42\textwidth}| p{0.54\textwidth}}
+\textit{Background question} & \textit{Subject responding YES}  \\ \hline
 \input{../analysis/background.tex}
-
  \end{tabular}
 \end{table}
+\endgroup
 
-\noindent \textbf{Frequency analysis.} The idea of ''essentiality'' is characterized as the number of subjects that has responded that they (1) use the concepts at least in an informal or non-persistent way, \textit{and} that they (2) use the concepts in a similar meaning as in the definition in Appendix A.
-Figure~\ref{fig:survey} shows the definitions of the three-level ordinal scales of Questions $Q1_{usage}$  and $Q2_{meaning}$ respectively.
+\vspace{-0.6cm}
+\noindent \textbf{Frequency analysis.} The degree of ''essentiality'' is characterized as the number of subjects that has responded that they (1) use the concept at least in an informal, non-persistent way, \textit{and} that they (2) use the concept in a similar meaning as in the definition in Appendix A.
+Fig.~\ref{fig:survey} shows the definitions of the three-level ordinal scales of Questions $Q1_{usage}$  and $Q2_{meaning}$ respectively.
 Table \ref{table:frequency} shows the results of the frequency counts.
-If an ''essentiality threshold'' is chosen at $N/2$ then only the $9$ concepts from row $n = 7$ and below in Table~\ref{table:frequency} are considered ''non-essential'', hence showing that more than 90\% of the metamodel concepts have a majority of the subjects that use and agree upon them.
-All concepts have at least one subject that use it and agree with the definition.
-
+If an ''essentiality threshold'' is chosen at $N/2$ then only the $9$ concepts from row $n = 7$ and below in Table~\ref{table:frequency} are considered ''non-essential'', hence showing that more than 90\% of the metamodel concepts have a majority of the subjects that use them and agree upon their definitions.
+Each concept have at least one subject that uses it and agrees with its definition.
+{The anonymized data and analysis scripts (developed using Scala and Apache POI) are available at: \footnotesize\url{https://github.com/bjornregnell/reqT-survey}}
+\vspace{-0.4cm}
 %
@@
 \fontsize{7}{8}\selectfont
-\caption{Frequency analysis, where $n$ is the number of subjects that for the respective concepts answered $(Q1_{usage}>= 1)$ \textit{and} $(Q2_{meaning} = 2)$.
+\caption{Frequency analysis, where $n$ is the number of subjects that for the respective concept answered $(Q1_{usage}>= 1)$ \textit{and} $(Q2_{meaning} = 2)$.
 In total there are 92 concepts (49 entities, 15 relations and 28 attributes).
@@
 It can be questioned if ''essentiallity'' of a set of RE concepts can be characterized by how many RE scholars that use them.
-However, if someone use a certain concept and wants to model it, then the metamodel of the applied modelling approach needs to have it, in order for that person not to find that the metamodel is lacking vital parts.
-The presented survey is a pilot investigation with two main contributions: (1) the survey instrument and the data collection and analysis approach is feasible, and (2) for more than 90\% of the metamodel concept there is a majority of the 14 participating RE scholars that use them and agree upon their definition.
+However, if someone use a certain concept and wants to model it, then the metamodel of the applied modelling approach needs to have it, in order not to risk that the person finds that the metamodel lacks vital parts.
+The presented survey is a pilot investigation with two main contributions: (1) the presented survey instrument, the data collection and analysis approach, which are shown to be feasible, and (2) the result that for more than 90\% of the reqT metamodel concept there is a majority of the 14 participating RE scholars that use them and agree upon their definitions.
 
 {\bf Limitations.
-} Due to the limited number of subjects and the high degree of homogeneity among subjects with respect to background, it is difficult to analyse and draw conclusions about potential differences in opinions between e.g.
+} The main threat to external validity is the limited number of subjects.
+Due to few subjects and the high degree of homogeneity among subjects with respect to background, it is difficult to analyse and draw conclusions e.g.
+about potential differences in opinions between e.g.
 teachers and developers.
-Some subjects needed more time and completed their survey offline during the coming days, which may give a variation in how careful the responses were considered.
+Some subjects needed more time and completed their survey offline during the coming days, which may give a variation in how carefully the responses were considered.
 %The survey was conducted by the author and inventor of reqT, in conjunction with a seminar and demo of reqT.
@@
 
-{\bf Further work.} When developing a metamodel it is interesting not just to ask if the concepts to include are essential, but also to pose the question if is the set of concepts is complete.
+{\bf Further work.} When developing a metamodel it is interesting not just to ask if the concepts to include are essential, but also to pose the question if the set of concepts is complete.
 If some essential concept is missing from some stakleholder's viewpoint, then the metamodel is not sufficient.
@@
 %A major objective of this research preview paper is to expose the latest version of reqT to the community of RE scholars and to invite discussions and contributions.
+
+\vspace{-0.2cm}
 

### 2015-10-02 src/script.scala
add markdown analysis

@@ -54,6 +54,7 @@ object Extract {  // ----------- Process the Matrix and output results ---------
   val TABLE_REL = OUT_DIR + "relation-defs.tex"
   val FILE_BACKGROUND = OUT_DIR + "background.tex"
   val FILE_ESSENTIAL = OUT_DIR + "essential.tex"
+  val FILE_ANALYSIS = OUT_DIR + "ANALYSIS.md"
   val NL = "\\\\"
 
   def toInt(s: String): Int = Try(s.toDouble.toInt).getOrElse(0) 
@@ -149,13 +150,20 @@ object Extract {  // ----------- Process the Matrix and output results ---------
           freqOf(conceptFilter, countAtLeastVerdict(1, 2)) foreach (p => println(p._1 + " " + p._2.mkString(" ")))
     }}
 
-    appendlnToFile(FILE_ESSENTIAL)(s"%%%%%%%%%%%%% Summary table of (use, agree) >= (1, 2)")
+    appendlnToFile(FILE_ESSENTIAL)("%%%%%%%%%%%%% Summary table of (use, agree) >= (1, 2)")
+    writelnToFile (FILE_ANALYSIS)("## Analysis of requirements engineering concept essentiality\n\n")
+    appendlnToFile(FILE_ANALYSIS)("| n | Entities | Attributes | Relations |")
+    appendlnToFile(FILE_ANALYSIS)("|--:|:---------|:-----------|:----------|")
+
     type CMap = Map[Int, Vector[String]]
     val ent:  CMap = freqOf(isEntity,    countAtLeastVerdict(1, 2)).toMap.withDefaultValue(Vector[String]())   
     val attr: CMap = freqOf(isAttribute, countAtLeastVerdict(1, 2)).toMap.withDefaultValue(Vector[String]())   
     val rel:  CMap = freqOf(isRelation,  countAtLeastVerdict(1, 2)).toMap.withDefaultValue(Vector[String]())  
     def out(i: Int, m: CMap) = s""" & \\texttt{${m(i).mkString(", ")}}"""
-    (14 to 0 by -1).foreach{i => appendlnToFile(FILE_ESSENTIAL)(s"$$$i$$" + out(i, ent) + out(i, attr) + out(i, rel) + NL + "\\hline" )} 
+    (14 to 0 by -1).foreach{i => 
[…]

### 2015-10-02 src/script.scala
fix markdown analysis

@@ -159,10 +159,11 @@ object Extract {  // ----------- Process the Matrix and output results ---------
     val ent:  CMap = freqOf(isEntity,    countAtLeastVerdict(1, 2)).toMap.withDefaultValue(Vector[String]())   
     val attr: CMap = freqOf(isAttribute, countAtLeastVerdict(1, 2)).toMap.withDefaultValue(Vector[String]())   
     val rel:  CMap = freqOf(isRelation,  countAtLeastVerdict(1, 2)).toMap.withDefaultValue(Vector[String]())  
-    def out(i: Int, m: CMap) = s""" & \\texttt{${m(i).mkString(", ")}}"""
+    def tex(i: Int, m: CMap) = s""" & \\texttt{${m(i).mkString(", ")}}"""
+    def plain(i: Int, m: CMap) = s"""${m(i).mkString(", ")}"""
     (14 to 0 by -1).foreach{i => 
-        appendlnToFile(FILE_ESSENTIAL)(s"$$$i$$" + out(i, ent) + out(i, attr) + out(i, rel) + NL + "\\hline" )
-        appendlnToFile(FILE_ANALYSIS)(s"""| $i | ${out(i, ent)} | ${out(i, attr)} | ${out(i, rel)} |\n""" )
+        appendlnToFile(FILE_ESSENTIAL)(s"$$$i$$" + tex(i, ent) + tex(i, attr) + tex(i, rel) + NL + "\\hline" )
+        appendlnToFile(FILE_ANALYSIS)(s"""| $i | ${plain(i, ent)} | ${plain(i, attr)} | ${plain(i, rel)} |""" )
     } 
   }
 

### 2015-12-22 paper/paper-refsq2016-final.tex
add new version to be camera ready

--- 2015-10-02 paper/paper-refsq2016-submission.tex
@@
 
-%\documentclass[runningheads,a4paper]{llncs}
 \documentclass[a4paper]{llncs}
+%\documentclass[a4paper]{llncs}
 
@@
 \title{What is essential? -- A pilot survey on views about the requirements metamodel of reqT.org}
-%reqT --  A Requirements Modeling Tool for Code Lovers
-% a short form should be given in case it is too long for the running head
-%\titlerunning{DRAFT MANUSCRIPT - Please do not cite. Submitted to REFSQ'13.}
+\titlerunning{DRAFT MANUSCRIPT -- Submitted to REFSQ'16.}
 
@@
 [{\bf Context \& motivation}] This research preview paper presents ongoing work on the metamodel of a free software requirements modeling tool called reqT that is developed in an educational context.
-The work aims to make an initial validation of a survey instrument that elicits views on the metamodel of the reqT tool, which seek to engage computer science students in Requirements Engineering (RE) through an open source requirements engineering DSL embedded in the Scala programming language.
+The work aims to make an initial validation of a survey instrument that elicits views on the metamodel of the reqT tool, which seek to engage computer science students in Requirements Engineering (RE) through an open source DSL embedded in the Scala programming language.
 [{\bf Question}] The research question is: Which RE concepts are essential to include in the metamodel for a requirements engineering tool in an educational context?
@@
 Model(
-  Component("apperance") has (
+  Component("appearance") has (
     VariationPoint("color") has (
@@
     Variant("green") requires Variant("square")),
-  Component("apperance") requires VariationPoint("shape"), 
-  App("free") has Component("apperance"),  
+  Component("appearance") requires VariationPoint("shape"), 
+  App("free") has Component("appearance"),  
   App("free") binds (VariationPoint("shape") binds Variant("round")),
-  App("premium") has Component("apperance"),  
+  App("premium") has Component("appearance"),  
   App("premium") binds ( 
@@
 \section{Methodology and Data Collection}
-In order to validate RE scholar's opinions of the metamodel, a survey instrument was developed including the 49 entities, 15 relations and 28 attributes of the reqT Version 3.0 metamodel.
-All the concepts are listed with their definitions in Appendix A.\footnote{The survey is available at \scriptsize\url{https://github.com/reqT/reqT/tree/3.0.x/survey}} These 92 concepts and definitions were gathered from various sources including the IREB Glossary \footnote{\scriptsize\url{https://www.ireb.org/en/cpre/cpre-glossary/}}, wikipedia, terminology from agile development, variability \cite{metzger2007variability} and goal modelling terminology, and the text book\cite{Lauesen2002} used in an RE course at Lund Univ.\cref{note:ets170}
-\vspace{-0.2cm}
+In order to validate RE scholar's opinions of the metamodel, a survey instrument was developed including the 49 entities, 15 relations and 28 attributes.
+All concepts and definitions are listed in Appendix A.
+\footnote{The survey is available at \scriptsize\url{https://github.com/reqT/reqT/tree/3.0.x/survey}}
+The concepts were gathered from various sources including the IREB Glossary \footnote{\scriptsize\url{https://www.ireb.org/en/cpre/cpre-glossary/}}, Wikipedia,  agile development, variability \cite{metzger2007variability} and goal modelling, and the text book\cite{Lauesen2002} used in an RE course at Lund Univ.\cref{note:ets170}
+\vspace{-0.5cm}
 \begin{figure}[h]
 \centering
-\includegraphics[width=\textwidth]{img/survey-screen-dump}
+\includegraphics[width=0.99\textwidth]{img/survey-screen-dump}
 \caption{A screen dump of a part of the survey instrument.}
@@
 \end{figure}
+\vspace{-0.5cm}
 
-\noindent The data collection for the presented pilot run was made during a Swedish national network meeting with academic RE scholars in the spring of 2015.
-The survey was filled in during the meeting using the participants' own laptops  in a downloadable spreadsheet as shown in Fig.~\ref{fig:survey}.
+\noindent The data collection was made during a Swedish national network meeting with academic RE scholars in spring 2015.
+The survey was filled in during the meeting using the participants' own laptops  in a spreadsheet shown in Fig.~\ref{fig:survey}.
 The subjects were given around 20 minutes to complete the survey.
-Most of the subjects handed in the survey via email directly after the session, while a few finished it offline and emailed their responses after the meeting.
+Most of the subjects handed in the survey via email directly after the session, while a few finished it after the meeting.
 \section{Data Analysis}
@@
 The analyzed\footnote{One subject answered NO on all background questions and was therefore excluded.} \input{../analysis/summary.tex} The response rate was 100\% after a reminder was emailed to one missing subject.
+%\vspace{-0.6cm}
+
+%\noindent 
+\textbf{Frequency analysis.} The degree of ''essentiality'' is characterized as the number of subjects that has responded that they (1) use the concept at least in an informal, non-persistent way, \textit{and} that they (2) use the concept in a similar meaning as in the definition in Appendix A.
+Fig.~\ref{fig:survey} shows the definitions of the three-level ordinal scales of Questions $Q1_{usage}$  and $Q2_{meaning}$ respectively.
+Table \ref{table:frequency} shows the results of the frequency counts.
+If an ''essentiality threshold'' is chosen at $N/2$ then only the $9$ concepts from row $n = 7$ and below in Table~\ref{table:frequency} are considered ''non-essential'', hence showing that more than 90\% of the metamodel concepts have a majority of the subjects that use them and agree upon their definitions.
+Each concept has at least one subject that uses it and agrees with its definition.
+Listing 1 shows concepts that subjects reported as important omissions, where 'or' was the only concept that had consensus among several subjects as being omitted.
+{The anonymised data and analysis scripts (developed using Scala and Apache POI) are available at: \footnotesize\url{https://github.com/bjornregnell/reqT-survey}}
+\vspace{-0.4cm}
 
@@
 
-\vspace{-0.6cm}
-\noindent \textbf{Frequency analysis.} The degree of ''essentiality'' is characterized as the number of subjects that has responded that they (1) use the concept at least in an informal, non-persistent way, \textit{and} that they (2) use the concept in a similar meaning as in the definition in Appendix A.
-Fig.~\ref{fig:survey} shows the definitions of the three-level ordinal scales of Questions $Q1_{usage}$  and $Q2_{meaning}$ respectively.
-Table \ref{table:frequency} shows the results of the frequency counts.
-If an ''essentiality threshold'' is chosen at $N/2$ then only the $9$ concepts from row $n = 7$ and below in Table~\ref{table:frequency} are considered ''non-essential'', hence showing that more than 90\% of the metamodel concepts have a majority of the subjects that use them and agree upon their definitions.
-Each concept have at least one subject that uses it and agrees with its definition.
-{The anonymized data and analysis scripts (developed using Scala and Apache POI) are available at: \footnotesize\url{https://github.com/bjornregnell/reqT-survey}}
-\vspace{-0.4cm}
-%
+\vspace{-3em}
 \begingroup
@@
 \begin{table}
-\centering
+%\centering
+\begin{center}
 \fontsize{7}{8}\selectfont
@@
 In total there are 92 concepts (49 entities, 15 relations and 28 attributes).
-The higher up in the table, the more ''essential''.}
+The higher up in the table, the more ''essential''.
+For $n = 0, 2, 3, 5$ the were no concepts with answers by that number of subjects.}
 \label{table:frequency}
@@
 \end{tabular}
+\end{center}
+
+\vspace{0.1cm}
+\fontsize{9}{10}\selectfont 
+\textbf{Listing 1. ''Missing Concepts'' according to subjects:}
+S01: or, 
+S02: bug, threshold,
+S04: role, problem, motivates, and, or, pattern, submodel,
+S06: plug-in, informalism, 
+S07: full sentence, 
+S09: satisfaction, satisfies, customer,
+S11: system-of-interest, verification, validation, S13: context. 
 \end{table}
@@
 
-It can be questioned if ''essentiallity'' of a set of RE concepts can be characterized by how many RE scholars that use them.
-However, if someone use a certain concept and wants to model it, then the metamodel of the applied modelling approach needs to have it, in order not to risk that the person finds that the metamodel lacks vital parts.
-The presented survey is a pilot investigation with two main contributions: (1) the presented survey instrument, the data collection and analysis approach, which are shown to be feasible, and (2) the result that for more than 90\% of the reqT metamodel concept there is a majority of the 14 participating RE scholars that use them and agree upon their definitions.
+The presented survey is a pilot investigation with two main contributions: (1) the presented survey instrument together with the data collection and analysis approach, which are shown to be feasible in the presented context, and (2) the pilot study result that for more than 90\% of the reqT metamodel concepts a majority of the 14 participating RE scholars claim to use them and agree upon their definitions.
+Only 7 subjects found in total 19 concepts ''missing'' that where important to them.
 
 {\bf Limitations.
-} The main threat to external validity is the limited number of subjects.
+} It can be questioned if ''essentiallity'' of a set of RE concepts can be characterized by how many RE scholars that use them.
+However, if someone use a certain concept and wants to model it, then the metamodel of the applied modelling approach needs to have it, in order not to risk that the person finds the metamodel lacking.
+A major threat to external validity is the limited number of subjects.
 Due to few subjects and the high degree of homogeneity among subjects with respect to background, it is difficult to analyse and draw conclusions e.g.
@@
 {\bf Further work.} When developing a metamodel it is interesting not just to ask if the concepts to include are essential, but also to pose the question if the set of concepts is complete.
-If some essential concept is missing from some stakleholder's viewpoint, then the metamodel is not sufficient.
+If some essential concept is missing from some stakeholder's viewpoint, then the metamodel is not sufficient.
 With more subjects participating in the presented RE metamodel survey, the analysis of answers to further questions on alternative terms and missing concepts will be enabled and beneficial to the further development of a comprehensive and complete, but not overloaded, RE metamodel.
@@
 
-\vspace{-0.2cm}
-
-\subsubsection*{Acknowledgments.} This work is partly funded by  VINNOVA within the EASE project.
+%\vspace{-0.2cm}
+\vspace{0.2cm}
+{\small\noindent\textbf{Acknowledgments.} Thanks to Tobias Kaufmann and Klaus Pohl for contributions to the variability model in Section 2.
+This work is partly funded by  VINNOVA within the EASE project.
+}
 
@@
 %\hspace{-0.9cm}
-\section*{Appendix A: Definitions of Metamodel Concepts}
+\section*{Appendix A: Definitions of Metamodel Concepts of reqT v3.0}
 \begin{minipage}{1\linewidth}

### 2016-03-14 paper/proceedings-version/essential.tex
add proceedings version

%%%%%%%%%%%%% Summary table of (use, agree) >= (1, 2)
$14$ & \texttt{Class, Component, UseCase, Variant} & \texttt{Comment, Example, Max, Min, Title} & \texttt{implements, verifies}\\\hline
$13$ & \texttt{Configuration, Data, Design, Event, Quality, Scenario, Stakeholder, System, Term} & \texttt{Code, Constraints, Cost, FileName, Probability, Profit, Spec, Why} & \texttt{excludes, interactsWith, is, relatesTo, requires}\\\hline
$12$ & \texttt{Actor, Domain, Feature, Function, Interface, Module, Relationship, Release, Req, Risk, Service, State, Task, Test} & \texttt{Benefit, Capacity, Frequency, Input, Order, Output, Prio, Text, Value} & \texttt{has, impacts}\\\hline
$11$ & \texttt{Idea, Label, Member, Meta, MockUp, Section, User} & \texttt{Image} & \texttt{precedes, superOf}\\\hline
$10$ & \texttt{Goal, Story} & \texttt{Expectation} & \texttt{}\\\hline
$9$ & \texttt{App, Issue, Target, WorkPackage} & \texttt{Damage} & \texttt{binds, helps}\\\hline
$8$ & \texttt{Item, Product, Resource, VariationPoint} & \texttt{} & \texttt{deprecates}\\\hline
$7$ & \texttt{Breakpoint, Screen} & \texttt{Status} & \texttt{}\\\hline
$6$ & \texttt{Barrier} & \texttt{Deprecated} & \texttt{hurts}\\\hline
%$5$ & \texttt{} & \texttt{} & \texttt{}\\\hline
$4$ & \texttt{Ticket} & \texttt{} & \texttt{}\\\hline
%$3$ & \texttt{} & \texttt{} & \texttt{}\\\hline
%$2$ & \texttt{} & \texttt{} & \texttt{}\\\hline
$1$ & \texttt{Epic} & \texttt{Gist} & \texttt{}\\\hline
%$0$ & \texttt{} & \texttt{} & \texttt{}\\\hline

### 2016-03-14 paper/proceedings-version/paper-refsq2016-final.tex
add proceedings version

--- 2015-12-22 paper/paper-refsq2016-final.tex
@@
 [{\bf Context \& motivation}] This research preview paper presents ongoing work on the metamodel of a free software requirements modeling tool called reqT that is developed in an educational context.
-The work aims to make an initial validation of a survey instrument that elicits views on the metamodel of the reqT tool, which seek to engage computer science students in Requirements Engineering (RE) through an open source DSL embedded in the Scala programming language.
+The work aims to make an initial validation of a survey instrument that elicits views on the metamodel of the reqT tool, which aims to engage computer science students in Requirements Engineering (RE) through an open source DSL embedded in the Scala programming language.
 [{\bf Question}] The research question is: Which RE concepts are essential to include in the metamodel for a requirements engineering tool in an educational context?
-[{\bf Principal ideas}] A survey instrument is developed with a list of 92 concepts (49 entities, 15 relations and 28 attributes) and a set of questions for each concept that elicit the respondents' views on the usage and interpretation of each concept.
+[{\bf Principal ideas}] A survey instrument is developed, with a list of 92 concepts (49 entities, 15 relations and 28 attributes) and a set of questions for each concept, to elicit the respondents' views on the usage and interpretation of each concept.
 [{\bf Contribution}] The survey is initially validated in a pilot study involving 14 Swedish RE scholars as subjects.
@@
 When teaching RE modelling we may ask ourselves: What are the \textit{essential} RE concepts that we should include in a taught metamodel for requirements?
-This paper investigates this questions in conjunction with the on-going work of developing a metamodel for reqT \cite{reqT}, an open source requirements engineering tool used in RE education  \cite{Regnell2013}.
+This paper investigates this questions in conjunction with the on-going work of developing a metamodel for \href{http://reqT.org}{reqT.org}, an open source requirements engineering tool used in RE education  \cite{Regnell2013}.
 A survey instrument is presented aiming to elicit the frequency of RE term usage and the degree of interpretation agreement.
@@
 A reqT model written in the above syntax is actually valid Scala code that, when executed, generates a data structure that can be traversed and manipulated using Scala scripts.
-Also visualisations can be generated using GraphViz export and export to HTML and spreadsheet formats.
+Visualisations can be generated by export to GraphViz.
+Export is also available to HTML and spreadsheet formats.
 
@@
 \textbf{Subject background.} The background questions in the survey regards the role of the subject, as shown in Table \ref{table:background}.
-The analyzed\footnote{One subject answered NO on all background questions and was therefore excluded.} \input{../analysis/summary.tex} The response rate was 100\% after a reminder was emailed to one missing subject.
+The analyzed\footnote{One subject answered NO on all background questions and was therefore excluded.} \input{summary.tex} The response rate was 100\% after a reminder was emailed to one missing subject.
 %\vspace{-0.6cm}
@@
 Each concept has at least one subject that uses it and agrees with its definition.
-Listing 1 shows concepts that subjects reported as important omissions, where 'or' was the only concept that had consensus among several subjects as being omitted.
+
+The following 19 concepts were reported ''missing'': 
+S01: \emph{or}, 
+S02: \emph{bug}, \emph{threshold},
+S04: \emph{role}, \emph{problem}, \emph{motivates}, \emph{and}, \emph{or}, \emph{pattern}, \emph{submodel},
+S06: \emph{plug-in}, \emph{informalism}, 
+S07: \emph{full sentence}, 
+S09: \emph{satisfaction}, \emph{satisfies}, \emph{customer},
+S11: \emph{system-of-interest}, \emph{verification}, \emph{validation}, S13: \emph{context}. 
+Thus, the  concept 'or' was the only concept that had consensus among several subjects (S01, S04) as considered ''missing''.
+
 {The anonymised data and analysis scripts (developed using Scala and Apache POI) are available at: \footnotesize\url{https://github.com/bjornregnell/reqT-survey}} 
-\vspace{-0.4cm}
+\vspace{-0.3cm}
 
@@
 \label{table:background}
+\vspace{-0.2cm}
 \begin{tabular}{p{0.42\textwidth}| p{0.54\textwidth}}
 \textit{Background question} & \textit{Subject responding YES}  \\ \hline
-\input{../analysis/background.tex}
+\input{background.tex}
  \end{tabular}
@@
 
-\vspace{-3em}
+\vspace{-3.6em}
 \begingroup
@@
 \textit{$n$} & \textit{Entities} & \textit{Attributes} & \textit{Relations} \\ \hline
-\input{../analysis/essential.tex}
+\input{essential.tex}
 \end{tabular}
@@
 
-\vspace{0.1cm}
-\fontsize{9}{10}\selectfont 
-\textbf{Listing 1. ''Missing Concepts'' according to subjects:}
-S01: or, 
-S02: bug, threshold,
-S04: role, problem, motivates, and, or, pattern, submodel,
-S06: plug-in, informalism, 
-S07: full sentence, 
-S09: satisfaction, satisfies, customer,
-S11: system-of-interest, verification, validation, S13: context. 
+%\vspace{0.1cm}
+%\fontsize{9}{10}\selectfont 
+%\textbf{Table 3.} $Q3$ \emph{Missing Concepts} according to subjects:
+%S01: or, 
+%S02: bug, threshold,
+%S04: role, problem, motivates, and, or, pattern, submodel,
+%S06: plug-in, informalism, 
+%S07: full sentence, 
+%S09: satisfaction, satisfies, customer,
+%S11: system-of-interest, verification, validation, S13: context. 
 \end{table}
 \endgroup
+
+%%%
 
@@
 
-The presented survey is a pilot investigation with two main contributions: (1) the presented survey instrument together with the data collection and analysis approach, which are shown to be feasible in the presented context, and (2) the pilot study result that for more than 90\% of the reqT metamodel concepts a majority of the 14 participating RE scholars claim to use them and agree upon their definitions.
-Only 7 subjects found in total 19 concepts ''missing'' that where important to them.
+The presented survey is a pilot investigation with two main contributions: (1) the survey instrument together with the data collection and analysis approach, which are shown to be feasible in the presented context, and (2) the pilot study results: for more than 90\% of the 92 reqT metamodel concepts a majority of the 14 participating RE scholars claim to use them and agree upon their definitions.
+Only 1 concept was considered missing by more than one subject, while in total 19 additional concepts were reported missing by some subject.
 
 {\bf Limitations.
-} It can be questioned if ''essentiallity'' of a set of RE concepts can be characterized by how many RE scholars that use them.
-However, if someone use a certain concept and wants to model it, then the metamodel of the applied modelling approach needs to have it, in order not to risk that the person finds the metamodel lacking.
+} It can be questioned if ''essentiallity'' of a set of RE concepts can be characterized by how many RE scholars that use them and agree upon their definition, but it can also be argued that concept usage in an educational context is interesting to investigate when developing a metamodel for an academic RE tool.
+%However, if someone use a certain concept and wants to model it, then the metamodel of the applied modelling approach needs to have it, in order not to risk that the person finds the metamodel lacking.
  A major threat to external validity is the limited number of subjects.
@@
 teachers and developers.
-Some subjects needed more time and completed their survey offline during the coming days, which may give a variation in how carefully the responses were considered.
-%The survey was conducted by the author and inventor of reqT, in conjunction with a seminar and demo of reqT.
-In order to avoid any positive bias due to advocacy in favour of the reqT metamodel, the survey was held prior to the seminar and demo.
-This in turn may introduce a threat of limited knowledge among subjects of the idea behind the modelling approach in reqT
+Some subjects needed more time and completed their survey offline, which may give a variation in how carefully the responses were considered.
+%The survey was conducted in conjunction with a seminar and demo of reqT. 
+%In order to avoid any positive bias, the survey was held prior to the seminar and demo. This in turn may introduce a threat of limited understanding of reqT.
 
@@
 \textbf{\textit{Entity}} & \textbf{\textit{Definition}}  \\
-\input{../analysis/entity-defs.tex}
+\input{entity-defs.tex}
  \end{tabular}
@@
 \textbf{\textit{Attribute}} & \textbf{\textit{Definition}}  \\ 
-\input{../analysis/attribute-defs.tex}
+\input{attribute-defs.tex}
  & \\
  \textbf{\textit{Relation}} & \textbf{\textit{Definition}}  \\ 
-\input{../analysis/relation-defs.tex}
+\input{relation-defs.tex}
 

### 2016-03-14 paper/proceedings/attribute-defs.tex
rename dirs

\texttt{Benefit}&A characterisation of a good or helpful result or effect (e.g. of a feature).\\
\texttt{Capacity}&The largest amount that can be held or contained (e.g. by a resource).\\
\texttt{Code}&A collection of (textual) computer instructions in some programming language, e.g. Scala. Short for source code.\\
\texttt{Comment}&A note that explains or discusses some entity.\\
\texttt{Constraints}&A collection of propositions that restrict the possible values of a set of variables.\\
\texttt{Cost}&The expenditure of something, such as time or effort, necessary for the implementation of an entity.\\
\texttt{Damage}&A characterisation of the negative consequences if some entity (e.g. a risk) occurs.\\
\texttt{Deprecated}&A description of why an entity should be avoided, often because it is superseded by another entity, as indicated by a 'deprecates' relation.\\
\texttt{Example}&A note that illustrates some entity by a  typical instance.\\
\texttt{Expectation}&The required output of a test in order to be counted as passed.\\
\texttt{FileName}&The name of a storage of serialized, persistent data.\\
\texttt{Frequency}&The rate of occurrence of some entity. \\
\texttt{Gist}&A short and simple description of an entity, e.g. a function or a test.\\
\texttt{Image}&(The name of) a picture of an entity.\\
\texttt{Input}&Data consumed by an entity, \\
\texttt{Max}&The maximum estimated or assigned (relative) value.\\
\texttt{Min}&The minimum estimated or assigned (relative) value.\\
\texttt{Order}&The ordinal number of an entity (1st, 2nd, ...).\\
\texttt{Output}&Data produced by an entity, e.g. a function or a test.\\
\texttt{Prio}&The level of importance of an entity. Short for priority.\\
\texttt{Probability}&The likelihood that something (e.g. a risk) occurs.\\
\texttt{Profit}&The gain or return of some entity, e.g. in monetary terms.\\
\texttt{Spec}&A (detailed) definition of an entity. Short for specification\\
\texttt{Status}&A level of refinement of an entity (e.g. a feature) in the development process. \\
\texttt{Text}&A sequence of words (in natural language).\\
\texttt{Title}&A general or descriptive heading.\\
\texttt{Value}&An amount. An estimate of worth.\\
\texttt{Why}&A description of intention. Rationale.\\

### 2016-03-14 paper/submitted/paper-refsq2016-final.tex
rename dirs

--- 2016-03-14 paper/proceedings-version/paper-refsq2016-final.tex
@@
 \textbf{Subject background.} The background questions in the survey regards the role of the subject, as shown in Table \ref{table:background}.
-The analyzed\footnote{One subject answered NO on all background questions and was therefore excluded.} \input{summary.tex} The response rate was 100\% after a reminder was emailed to one missing subject.
+The analyzed\footnote{One subject answered NO on all background questions and was therefore excluded.} \input{../analysis/summary.tex} The response rate was 100\% after a reminder was emailed to one missing subject.
 %\vspace{-0.6cm}
@@
 Each concept has at least one subject that uses it and agrees with its definition.
-
-The following 19 concepts were reported ''missing'': 
-S01: \emph{or}, 
-S02: \emph{bug}, \emph{threshold},
-S04: \emph{role}, \emph{problem}, \emph{motivates}, \emph{and}, \emph{or}, \emph{pattern}, \emph{submodel},
-S06: \emph{plug-in}, \emph{informalism}, 
-S07: \emph{full sentence}, 
-S09: \emph{satisfaction}, \emph{satisfies}, \emph{customer},
-S11: \emph{system-of-interest}, \emph{verification}, \emph{validation}, S13: \emph{context}. 
-Thus, the  concept 'or' was the only concept that had consensus among several subjects (S01, S04) as considered ''missing''.
-
+Table~3 shows 19 concepts that subjects reported as ''missing'' ($Q3$), where 'or' was the only concept that had consensus among several subjects as considered ''missing''.
 {The anonymised data and analysis scripts (developed using Scala and Apache POI) are available at: \footnotesize\url{https://github.com/bjornregnell/reqT-survey}}
@@
 \textit{Background question} & \textit{Subject responding YES}  \\ \hline
-\input{background.tex}
+\input{../analysis/background.tex}
  \end{tabular}
@@
 
-\vspace{-3.6em}
+\vspace{-3.2em}
 \begingroup
@@
 \textit{$n$} & \textit{Entities} & \textit{Attributes} & \textit{Relations} \\ \hline
-\input{essential.tex}
+\input{../analysis/essential.tex}
 \end{tabular}
@@
 
-%\vspace{0.1cm}
-%\fontsize{9}{10}\selectfont 
-%\textbf{Table 3.} $Q3$ \emph{Missing Concepts} according to subjects:
-%S01: or, 
-%S02: bug, threshold,
-%S04: role, problem, motivates, and, or, pattern, submodel,
-%S06: plug-in, informalism, 
-%S07: full sentence, 
-%S09: satisfaction, satisfies, customer,
-%S11: system-of-interest, verification, validation, S13: context. 
+\vspace{0.1cm}
+\fontsize{9}{10}\selectfont 
+\textbf{Table 3.} $Q3$ \emph{Missing Concepts} according to subjects:
+S01: or, 
+S02: bug, threshold,
+S04: role, problem, motivates, and, or, pattern, submodel,
+S06: plug-in, informalism, 
+S07: full sentence, 
+S09: satisfaction, satisfies, customer,
+S11: system-of-interest, verification, validation, S13: context. 
 \end{table}
@@
 \textbf{\textit{Entity}} & \textbf{\textit{Definition}}  \\
-\input{entity-defs.tex}
+\input{../analysis/entity-defs.tex}
  \end{tabular}
@@
 \textbf{\textit{Attribute}} & \textbf{\textit{Definition}}  \\ 
-\input{attribute-defs.tex}
+\input{../analysis/attribute-defs.tex}
  & \\
  \textbf{\textit{Relation}} & \textbf{\textit{Definition}}  \\ 
-\input{relation-defs.tex}
+\input{../analysis/relation-defs.tex}
 

### 2016-03-14 paper/presentation/presentation.tex
add presentation

\documentclass{slides}

\setbeamertemplate{footline}[frame number]
\title[What is essential?]{What is essential?}
\subtitle{\footnotesize -- A pilot survey on views about the requirements metamodel of reqT.org}
\author{Björn Regnell}
\institute{Lund University}
\date{March 14, 2016\\ \href{http://refsq.org/2016}{refsq.org/2016}\\ \href{https://github.com/bjornregnell/reqT-survey}{github.com/bjornregnell/reqT-survey}}

\begin{document}

\frame{\titlepage}
\frame{\tableofcontents}

\section{Objective}
\subsection{Research Question}
\begin{Slide}{Research question}
In the context of software requirements engineering education:
\begin{itemize}
\item How to choose a set of \\ \Emph{essential requirements engineering concepts} \\ that allows for sufficient expressiveness, \\ without overloading the metamodel with esoteric concepts just for the sake of completeness?
\end{itemize}
\end{Slide}
\subsection{Approach}
\begin{Slide}{Approach}
Make a survey among RE scholars:
\begin{itemize}
\item How to quantify ''essentiality''?
\pause
\item One possible quantification: \\ The more scholars that agree on a definition of a concept \\ and \\ the more scholars that use the concept, \\ the more essential is the concept.
\end{itemize}
\end{Slide}

\end{document}

### 2016-03-14 paper/presentation/presentation.tex
add some color

@@
 \begin{itemize}
-\item How to choose a set of \\ \Emph{essential requirements engineering concepts} \\ that allows for sufficient expressiveness, \\ without overloading the metamodel with esoteric concepts just for the sake of completeness?
+\item How to choose a set of \\ \Emph{essential requirements engineering concepts} \\ that allows for \Alert{sufficient expressiveness}, \\ without overloading the metamodel with esoteric concepts just for the sake of \Alert{completeness}?
 \end{itemize}
@@
 \pause
-\item One possible quantification: \\ The more scholars that agree on a definition of a concept \\ and \\ the more scholars that use the concept, \\ the more essential is the concept.
+\item One possible quantification: \\ The more scholars that \Emph{agree} on a definition of a concept \\ and \\ the more scholars that \Emph{use} the concept, \\ the more essential is the concept.
 \end{itemize}

### 2016-03-15 paper/presentation/discussant.tex
add a boat load of stuff

\documentclass{slides}

\setbeamertemplate{footline}[frame number]
\title{REFSQ2016}
\begin{document}

\section{Paper discussion in session 5a: RE Foundations}
\subsection{Foundations for Transparency Requirements Engineering, Hosseini et al.}

\begin{Slide}{Discussion items (Björn Regnell)}
\begin{itemize}
\item  \Emph{What can we do now that we could not do before?} \\ \pause
  Apply 4 facets of transparency in our RE practice
\pause\item  \Emph{How sound is the solution?} \\ \pause
  Based on a literature review, ''extensive'', systematic?
\pause\item    \Emph{What is the next step to take?} \\ \pause
        Test it in an empirical case study using real-world data
\pause\item    \Emph{What is the most questionable issue of the paper?} \\ \pause
  \Alert{The???} 4 facets of transparency
\pause
\begin{itemize}
\item What have you missed? (Completeness)
\item What is essential? Apply ''Occam's razor''
\item What is your moral stand-point on transparency?
\end{itemize}
\end{itemize}
\end{Slide}

\end{document}

### 2016-03-15 paper/presentation/presentation.tex
add slides
add stuff
add a boat load of stuff

@@
 \date{March 14, 2016\\ \href{http://refsq.org/2016}{refsq.org/2016}\\ \href{https://github.com/bjornregnell/reqT-survey}{github.com/bjornregnell/reqT-survey}}
+
+\usepackage{xcolor}
+\definecolor{entityColor}{RGB}{0,100,200}
+\definecolor{attributeColor}{RGB}{0,100,50}
+\definecolor{relationColor}{RGB}{160,0,30}
+
+\lstdefinestyle{reqT}{
+  %belowcaptionskip=1\baselineskip,
+  breaklines=true,
+  %showstringspaces=false,
+  showspaces=false,
+  %breakatwhitespace=true,
+  basicstyle=\ttfamily\fontsize{8}{10}\selectfont,
+  emph={Ent,Meta,Item,Label,Section,Term,Actor,App,Component,Domain,Module,Product,Release,Resource,Risk,Service,Stakeholder,System,User,Class,Data,Input,Member,Output,Relationship,Design,Screen,MockUp,Function,Interface,State,Event,Epic,Feature,Goal,Idea,Issue,Req,Ticket,WorkPackage,Breakpoint,Barrier,Quality,Target,Scenario,Task,Test,Story,UseCase,VariationPoint,Variant},
+  emphstyle=\bfseries\color{entityColor},
+  emph={[2]has,is,superOf,binds,deprecates,excludes,helps,hurts,impacts,implements,interactsWith,precedes,requires,relatesTo,verifies},
+  emphstyle={[2]\color{relationColor}},
+  emph={[3]Attr,Code,Constraints,Comment,Deprecated,Example,Expectation,FileName,Gist,Image,Spec,Text,Title,Why,Benefit,Capacity,Cost,Damage,Frequency,Min,Max,Order,Prio,Probability,Profit,Value,Status},
+  emphstyle={[3]\itshape \color{attributeColor}},  
+}
+\lstset{style=reqT, language=}
 
@@
 \item How to choose a set of \\ \Emph{essential requirements engineering concepts} \\ that allows for \Alert{sufficient expressiveness}, \\ without overloading the metamodel with esoteric concepts just for the sake of \Alert{completeness}?
+\pause
+\item Presumption: as teachers we should be method agnostic; there is no single ''correct'' dogma
 \end{itemize}
@@
 \begin{Slide}{Approach}
-Make a survey among RE scholars:
+\begin{itemize}
+
+\item Make a survey among RE scholars
+\pause
 \begin{itemize}
@@
 \pause
-\item One possible quantification: \\ The more scholars that \Emph{agree} on a definition of a concept \\ and \\ the more scholars that \Emph{use} the concept, \\ the more essential is the concept.
+\begin{itemize}
+\item One possible quantification: \\ The more scholars that \Emph{agree} on a definition of a concept \\ and \\ the more scholars that \Emph{use} the concept, \\ the more \Alert{essential} is the concept.
+\end{itemize}
+\end{itemize}
+\pause
+\item Use the reqT metamodel as a basis for the survey
 \end{itemize}
@@
 
+\section{Background}
+\subsection{About reqT}
+\begin{Slide}{What is reqT?}
+\begin{itemize}
+\item reqT is an open source requirements engineering tool developed in an \Emph{educational} context\\ Download from here: \url{http://reqt.org/} \\ \texttt{java -jar reqT.jar}
+\item reqT aims to inspire \Alert{code-loving} cs students to learn more about RE through a \Emph{Scala}-internal DSL
+\item reqT has a \Emph{metamodel} including entities, relations and attributes gathered from various sources such as text books, research papers, IREB, wikipedia, etc.
+\end{itemize}
+\end{Slide}
+
+\begin{Slide}{A small reqT model}
+\begin{Code}
+val m = Model(
+  Feature("chat") has (
+    Spec("The system shall enable sending text messages among users"),
+    Prio(42)
+  )
+)
+\end{Code}
+\end{Slide}
+
+\begin{Slide}{An larger example reqT model}
+\begin{Code}
+Model(
+  Component("appearance") has (
+    VariationPoint("color") has (
+      Min(0), Max(2), Variant("blue"), Variant("red"), Variant("green")),
+    VariationPoint("shape") has (
+      Min(1), Max(1), Variant("round"), Variant("square")),
+    VariationPoint("payment") has (
+      Min(1), Max(2), Variant("cash"), Variant("credit")),
+    VariationPoint("payment") requires Variant("cash"), 
+    Variant("round") excludes Variant("red"),
+    Variant("green") requires Variant("square")),
+  Component("appearance") requires VariationPoint("shape"), 
+  App("free") has Component("appearance"),  
+  App("free") binds (VariationPoint("shape") binds Variant("round")),
+  App("premium") has Component("appearance"),  
+  App("premium") binds ( 
+    VariationPoint("color") binds (Variant("red"), Variant("green")),
+    VariationPoint("shape") binds (Variant("round"), Variant("square")),
+    VariationPoint("payment") binds Variant("cash")))
+\end{Code}
+\end{Slide}
+
+\begin{Slide}{49 entities of the metamodel (see paper appendix)}
+\fontsize{3.5}{4.0}\selectfont
+\begin{tabular}{l l}
+\textbf{\textit{Entity}} & \textbf{\textit{Definition}}  \\
+\input{../proceedings/entity-defs.tex}
+\end{tabular}
+\end{Slide}
+
+\begin{Slide}{28 atributes of the metamodel (see paper appendix)}
+\fontsize{6}{6.5}\selectfont
+\begin{tabular}{l l}
+\textbf{\textit{Attributes}} & \textbf{\textit{Definition}}  \\
+\input{../proceedings/attribute-defs.tex}
+\end{tabular}
+\end{Slide}
+
+\begin{Slide}{15 realtions of the metamodel (see paper appendix)}
+\fontsize{8}{9}\selectfont
+\begin{tabular}{l l}
+\textbf{\textit{Relation}} & \textbf{\textit{Definition}}  \\
+\input{../proceedings/relation-defs.tex}
+\end{tabular}
+\end{Slide}
+
+\section{Methodology}
+\subsection{Data Collection}
+\begin{Slide}{Data Collection}
+{\centering
+\includegraphics[width=0.8\textwidth]{img/survey-screen-dump}
+}
+Q1 use? =   no, orally, also in writing\\
+Q2 agree? = no, don't know, yes  \\
+Answered by 15 swedish RE scholars (100\% response rate)\\
+\url{https://github.com/bjornregnell/reqT-survey}
+
+\end{Slide}
+
+\subsection{Data Analysis}
+\begin{Slide}{Data Analysis}
+\end{Slide}
+
+\section{Results}
+\subsection{Essentiality}
+\begin{Slide}{Essentiality}
+\end{Slide}
+
+\section{Conclusion}
+\subsection{Future Work}
+\begin{Slide}{Future Work}
+\end{Slide}
+
 \end{document}

### 2016-03-15 paper/presentation/presentation.tex
add more stuff

@@
 \setbeamertemplate{footline}[frame number]
-\title[What is essential?]{What is essential?}
+\title[What is essential? -- A research preview]{What is essential?}
 \subtitle{\footnotesize -- A pilot survey on views about the requirements metamodel of reqT.org}
 \author{Björn Regnell}
-\institute{Lund University}
-\date{March 14, 2016\\ \href{http://refsq.org/2016}{refsq.org/2016}\\ \href{https://github.com/bjornregnell/reqT-survey}{github.com/bjornregnell/reqT-survey}}
+\institute{Lund University, Sweden}
+\date{\href{http://refsq.org/2016}{refsq.org/2016}\\ \href{https://github.com/bjornregnell/reqT-survey}{github.com/bjornregnell/reqT-survey}}
 
@@
 \pause
-\item Use the reqT metamodel as a basis for the survey
+\item Use the reqT.org metamodel as a basis for the survey
 \end{itemize}
@@
 \item reqT is an open source requirements engineering tool developed in an \Emph{educational} context\\ Download from here: \url{http://reqt.org/} \\ \texttt{java -jar reqT.jar}
-\item reqT aims to inspire \Alert{code-loving} cs students to learn more about RE through a \Emph{Scala}-internal DSL
-\item reqT has a \Emph{metamodel} including entities, relations and attributes gathered from various sources such as text books, research papers, IREB, wikipedia, etc.
+\item reqT aims to inspire \Emph{code-loving} cs students to learn more about RE through a \Emph{Scala}-internal DSL
+\item has a metamodel with {\color{entityColor}{\bf entities}}, {\color{relationColor}{\bf relations}} and {\color{attributeColor}{\bf attributes}}\\ gathered from various sources such as \\ text books, research papers, IREB, wikipedia, etc.
 \end{itemize}
@@
 }
-Q1 use? =   no, orally, also in writing\\
-Q2 agree? = no, don't know, yes  \\
+{\small Q1 use =   \{no | used, but only orally | used, also in writing\}\\ Q2 agree = \{no, different meaning | don't know | yes, similar meaning\}  \\
 Answered by 15 swedish RE scholars (100\% response rate)\\
 \url{https://github.com/bjornregnell/reqT-survey}
+}
 
@@
 \begin{Slide}{Data Analysis}
+\textbf{Frequency analysis.} The degree of ''essentiality'' is characterized as the number of subjects that has responded that they (1) use the concept at least in an informal, non-persistent way, \textit{and} that they (2) use the concept in a similar meaning as in the definition in Appendix A.
+
 \end{Slide}
@@
 \subsection{Essentiality}
-\begin{Slide}{Essentiality}
+\begin{Slide}{Frequency Analysis}
+\fontsize{6}{8}\selectfont
+
+\begin{tabular}{l | p{0.33\textwidth} | p{0.30\textwidth} | p{0.25\textwidth}}
+\textit{$n$} & \textit{Entities} & \textit{Attributes} & \textit{Relations} \\ \hline
+\input{../proceedings/essential.tex}
+\end{tabular}
+
+\end{Slide}
+
+\begin{Slide}{Essentiality Definition}
+\begin{itemize}
+\item If an ''essentiality threshold'' is chosen at $N/2$ then \\ only the $9$ concepts from row $n = 7$ and below are considered ''non-essential''
+\item \Alert{More than 90\% of the 92 metamodel concepts have a majority of the subjects that use them and agree upon their definitions} 
+\item Each concept has at least one subject that uses it and agrees with its definition. 
+\end{itemize}
+
 \end{Slide}
@@
 \section{Conclusion}
-\subsection{Future Work}
-\begin{Slide}{Future Work}
+\subsection{Contribution \& Future Work}
+\begin{Slide}{Conlusion}
+
+\begin{itemize} 
+\item Contribution
+\begin{itemize} 
+\item A way to quantify ''essentiality'' -- tested in a pilot survey
+\item A set of potentially ''essential'' concepts \\ (according to 14 RE scolars in Sweden)
+\end{itemize}
+\end{itemize}
+
+\begin{itemize} 
+\item Future Work
+\begin{itemize} 
+\item More subjects for increased external validity
+\item More questions on missing and alternative concepts
+\end{itemize}
+\end{itemize}
 \end{Slide}

### 2016-03-17 paper/presentation/presentation.tex
add metamodel
add stuff

@@
 
+\begin{Slide}{The base classes of the reqT metamodel}
+\includegraphics[width=0.9\textwidth]{img/metamodel}
+\end{Slide}
+
 \begin{Slide}{A small reqT model}
@@
 
-\begin{Slide}{28 atributes of the metamodel (see paper appendix)}
+\begin{Slide}{28 attributes of the metamodel (see paper appendix)}
 \fontsize{6}{6.5}\selectfont
@@
 \end{itemize}
+\end{Slide}
+
+\begin{Slide}{Omissions}
+%\vspace{0.1cm}
+%\fontsize{9}{10}\selectfont 
+\Alert{Missing Concepts} according to subjects:\\
+S01: \Emph{or}, \\
+S02: bug, threshold, \\
+S04: role, problem, motivates, and, \Emph{or}, pattern, submodel, \\
+S06: plug-in, informalism, \\
+S07: full sentence, \\
+S09: satisfaction, satisfies, customer, \\
+S11: system-of-interest, verification, validation, \\
+S13: context.
 
@@
 \begin{itemize} 
-\item More subjects for increased external validity
-\item More questions on missing and alternative concepts
+\item Increase validity
+\item Investigate omissions
 \end{itemize}
